---
title: "curvana: a comprehensive R package for analyzing atomic force microscopy force distance curves"
subtitle: "Tutorial"
date: "`r Sys.Date()`"
author: Yue (Gary) Zhang, PhD
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
editor_options: 
  chunk_output_type: console
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(warning = T, message = T,echo = T) 
```

```{r Initiate, results='hide'}
## initiate ---------------------------------------------------------
packages_to_load = c('curvana','ggplot2','DT','dplyr','DT')
# Load packages without printing output
invisible(suppressPackageStartupMessages({
    lapply(packages_to_load, library, character.only = TRUE)
}))

```

# Overview

**curvana** is an R package that provides a comprehensive framework for analyzing
atomic force microscopy (AFM) force–distance curves.  
It is built around an S4 object class called **`fdObj`**, which standardizes
data import, storage, extraction, and analysis.  

The `fdObj` structure organizes all raw and processed data into designated slots,
ensuring data integrity, simplifying analysis workflows, and facilitating
reproducible data sharing.

# Core Workflow

**curvana** accepts as input a directory of raw *deflection–displacement* curves.
All curves are automatically registered in the internal **metadata** slot, while
the corresponding raw data are stored in the **rawcurves** slot.

Once imported, users can apply curvana’s analysis functions to:

- **Transform** raw deflection–displacement curves into calibrated **force–distance** curves  
- **Quantify** key mechanical properties such as:
  - Adhesive force  
  - Adhesive and repulsive energies  
  - Rupture and repulsive distances  

All computed results are automatically saved back into the `fdObj` metadata,
maintaining a complete, self-contained record of the analysis.

# Visualization and Reporting

curvana includes flexible plotting utilities for generating publication-ready figures:

- **Scatter plots** of individual or batch-wise raw and processed curves  
- **Violin plots** summarizing distributions of curve-derived properties  

These tools make it easy to explore data quality, compare experimental conditions,
and visualize quantitative trends across samples.

# Integration with shinyCurvana

**curvana** seamlessly integrates with **shinyCurvana**, an R Shiny application that
provides a **code-free, user-friendly interface** for performing the same analyses
interactively.  

Together, they enable both scripting-based and graphical workflows for AFM
force–distance data processing and interpretation.

```{r save image}
save.image(file = "")
```
