---
title: "curvana: a comprehensive R package for analyzing atomic force microscopy force–distance curves"
subtitle: "Tutorial"
author: Yue (Gary) Zhang, PhD
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    number_sections: false
    highlight: tango
vignette: >
  %\VignetteIndexEntry{curvana tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=TRUE}
knitr::opts_chunk$set(warning = T, message = T,echo = T) 
```


# Introduction

This tutorial demonstrates the full analytical workflow of the **curvana** R package, from raw AFM deflection–displacement curves to quantitative force–distance curves and analytical metrics, such as adhesive force, interaction energy and distance.  

# Preparation 

## Initializing work environment

```{r Initiate, results='hide',warning='False'}
packages_to_load = c('curvana','ggplot2','DT','dplyr','DT','ggpubr','cowplot')
invisible(suppressPackageStartupMessages({
    lapply(packages_to_load, library, character.only = TRUE)
}))
source(system.file("tutorial_helpers.R", package = "curvana"))
```

## Load demo data

All examples below use the demo dataset provided in `extdata/` inside the package. To use your own dataset, you can pass the path to the directory holding the data to the "folder" argument. Note that users can increase the threads to enable parallel processing for loading a large folder (e.g., > 100 curves), which will dramatically reduce the computational time. Here we use 1 thread, as we only have a small demo dataset.

At this step, we initialize the fdObj named "test" by importing the demo data directory "extdata/" using the "createFdObjFromFolder" function. An fdObj is the core data container in curvana, designed to store atomic force microscopy (AFM) force–distance curve data, including raw approach and retract curves, transformed data, associated metadata for downstream analysis, and the analytical metrics generated.

By default, createFdObjFromFolder will read all files in the designated folder with suffix ".txt", and expect each file has paired approach and retract deflection-dislocation raw curve data with standard column names (i.e., Calc_Ramp_Ex_nm, Defl_V_Ex, Calc_Ramp_Rt_nm, Defl_V_Rt). 

Users working with non-standard input files can specify column headers with corresponding arguments. User may also use the "suffix" argument to specify files to import by specifying filename suffix, or by providing a metadata dataframe whose row names match the input file names. Please note that the input files have to be tab-delimited files. See package manual for detailed usage.

```{r load data}
folder = system.file("extdata", package = "curvana")
t = 1
test = createFdObjFromFolder(folder = folder, threads = t)
```

Let's take a quick look at the fdObj (named test) we created. We can see that we imported 14 raw curves. Non of them are transformed yet, and "filename" is the only column in the metadata slot.

```{r show init fdobj}
print(test)
view_df(test@metadata,'metadata')
```

## Manipulate the fdObj object

The demo data set has measurements made on different bacterial biofilm surface at different dates. Let's group them by setting the date and bacteria group columns

```{r manipulate fdobj by separating data into groups}
test@metadata$date = sub(pattern = '_.*','',test@metadata$filename)
test@metadata$bacteria = sub(".*f?_(.*)_loc.*", "\\1", test@metadata$filename)
view_df(test@metadata,'metadata')
```

It is also possible to extract specific samples from an fdObj either by providing their sample names (i.e., row names in the metadata) or by subsetting the object based on conditions applied to metadata columns.

```{r manipulate fdobj with extract}
test.paeru = extract(test,by_col = list(bacteria = "paeru"))
test.paeru2 = extract(test,by_sample = rownames(test@metadata[test@metadata$bacteria == 'paeru',]))
test.pmon = extract(test,by_col = list(bacteria = 'pmon2'))
```

Just as we might need to subset a fdObj, we might also need to combine two fdObjs (e.g., from different experiments). The "combine" function is designed for this purpose. Note that if there are overlapping samples between the two fdObjs being combined, the "combine" function will return an error.

```{r manipulate fdobj with combine}
test.combine.1 = combineFdObj(test.paeru,test.pmon)
#test.combine.2 = combineFdObj(test.paeru,test.paeru2) # should return error
```

## Preview the raw deflection curves

It is important to inspect the raw curves before formal analysis. We can use the "plot_deflection_curves" function to plot curves in batch. Here, we split curves into sub-panels by column "bacteria", and color code them by column "date".

```{r plot deflection curves, fig.height=7,fig.width=7}
plot_deflection_curves(test,split_curves_by = 'bacteria',group_curves_by = 'date',
                       alpha = 0.5,point_size = 0.5,
                       color_map = list(`180424` = 'darkred',
                                        `180425` = 'orange'))
```

# Curve transformation

One advantage of curvana is that each step of the data analysis is modularized, enabling a flexible and customizable workflow. Curve transformation in curvana involves several steps, including sensitivity calculation, baseline offset calculation, and the application of these parameters to convert raw data into interaction force–separation distance curves. In this section, we demonstrate how to apply a wrapper function to transform all curves at once, as well as how each curve is transformed internally under the hood.

## Transform curves using the wrapper function "transform_curves"

At this step, we calculate the baseline and sensitivity of each raw AFM curve and transform the deflection-dislocation data into calibrated force–distance curves. The wrapper function transform_curves() automates this process by internally calling analyze_sensitivity() and analyze_baseline() to estimate the sensitivity and baseline values when their corresponding columns are not yet present in the metadata.

To transform a curve, the spring constant of the AFM probe is required to convert deflection (V) to force (nN). The transform_curves() function accepts either a single numeric value, applied to all curves and stored in metadata$spring_constant, or a metadata column name containing per-curve spring constants. These values are then used with the computed baseline and sensitivity to generate the final force–distance curves.

After these parameters are obtained, the function applies them to transform each curve into its calibrated form. All arguments accepted by analyze_sensitivity() and analyze_baseline() can be passed directly to transform_curves(), allowing full control over the analysis parameters.

```{r transform curves approach,warning=TRUE, message=TRUE,results='hold', echo=TRUE}
t = 1
test = transform_curves(test,spring_constant = 0.3188,useCurve = 'approach',threads = t)
```

```{r transform curves retract,results='hold'}
t = 1
test = transform_curves(test,spring_constant = 0.3188,useCurve = 'retract',threads = t)
```

## Examine the object after transformation

We can see that now we have curves transformed.

```{r examine fdobj after transformation}
print(test)
```

The transformed curves are stored in the retractCurves and approachCurves slots of the fdObj. Note that when the sensitivity or baseline cannot be reliably determined—such as in curves with highly fluctuating probe–surface contact regions or irregular baseline segments—the transformation will fail. In these cases, the corresponding entry in the transformed curve slot (retractCurves or approachCurves) is stored as an empty data frame with zero rows, indicating an unsuccessful transformation.

```{r show retract and approach curve slots}
print(names(test@retractCurves))
print(names(test@approachCurves))
```

The calculated sensitivity, baseline, and input spring constant is now saved as columns in the metadata. Note the curves with baseline/sensitivity undetermined will be NA in the corresponding cell.

```{r show metadata after transformation}
view_df(test@metadata,'metadata after curve transformation')
```


## Single curve transformation

### Sensitivity calculation

To demonstrate how each single curve is transformed, we take the approach part of "180424_fc_mica_loc1.001" as an example. The function "calc_sensitivity", adapted from fdafm (Zhang et al., 2021), is designed to automatically to calculate the sensitivity value. Note that the function returns two outputs -- the sensitivity value, and the subset of data points used to calculate the sensitivity value (the second element of the returned list, which is a named dataframe named senscal_segment).

```{r single curve sensitivity analysis}
sc.1.name = '180424_fc_mica_loc1.001'
sc.1.raw = test@rawCurves[[sc.1.name]]

x = sc.1.raw$Calc_Ramp_Ex_nm
y = sc.1.raw$Defl_V_Ex
res = calc_sensitivity(end = 200,intv = 4,x = x,y = y)
sc.1.raw$`Region used for Sensitivity calculation` = "FALSE"
sc.1.raw$`Region used for Sensitivity calculation`[1:nrow(res$senscal_segment)] <- 'TRUE'

ggplot(sc.1.raw, aes(x = Calc_Ramp_Ex_nm, y = Defl_V_Ex)) +
  geom_point(aes(
    color = case_when(
      `Region used for Sensitivity calculation` == TRUE ~ "Sensitivity region",
      TRUE ~ "Other"
    )
  ), size = 1.5) +
  scale_color_manual(
    values = c(
      "Sensitivity region" = "hotpink",
      "Baseline region" = "orange",
      "Other" = "grey70"
    ),
    name = "Region"
  ) +
  labs(
    x = "Calc_Ramp_Ex (nm)",
    y = "Defl_V_Ex (V)",
    title = "AFM Force Curve Regions"
  ) +
  theme_minimal(base_size = 14)
```

To analyze the sensitivity value for all curves in the object at once, we can run function "analyze_sensitivity." The function internally executes "calc_sensitivity" for each curve separately, and record the sensitivity value of each curve in a metadata column named "sensitivity_V_nm_dir", where dir is the selected curve portion via useCurve argument (either "approach" or "retract"). 

Curves that failed in sensitivity calculation will have an 'NA' in their corresponding cells. The segment of each raw curve used for sensitivity calculation is saved in the object's slot named "senscal_segment".

```{r analyze sensitivity in batch}
test = analyze_sensitivity(test,useCurve = 'approach',threads = 1)
test = analyze_sensitivity(test,useCurve = 'retract',threads = 1)
view_df(test@metadata,'metadata with sensitivity values added as columns')
```

### Baseline offsets

Similarly, the offset voltage of each curve can be estimated using the find_baseline() function. We assume that the far-right region of every curve corresponds to the probe being fully detached from the substrate and therefore free from interaction forces. The mean of the first N points (defined by the least_length argument, default = 150) in this region is calculated as the baseline offset value.

Some curves exhibit undulating or unstable baselines. Such abnormal curves—identified by a slope greater than slp_threshold (default = 0.001) and a standard deviation greater than std_threshold (default = 0.005)—are considered unreliable and are omitted from subsequent analyses.

Similar as the "cal_sensitivity" function, "find_baseline" also returns two outputs -- the offset value, and the subset of data points used to calculate the offset value (the second element of the returned list, which is a named dataframe named baseline_segment).


```{r single curve baseline calculation}

res.b = find_baseline(x = x, y = y, least_length = 150, sensitivity = res$sensitivity,slp_threshold = 0.001,std_threshold = 0.005)
sc.1.raw$`Region used for baseline offset calculation` = 'TRUE'
sc.1.raw$`Region used for baseline offset calculation`[1:(length(sc.1.raw$Defl_V_Ex)-150)] = 'FALSE'

ggplot(sc.1.raw, aes(x = Calc_Ramp_Ex_nm, y = Defl_V_Ex)) +
  geom_point(aes(
    color = case_when(
      `Region used for baseline offset calculation` == TRUE ~ "Baseline region",
      TRUE ~ "Other"
    )
  ), size = 1.5) +
  scale_color_manual(
    values = c(
      "Sensitivity region" = "hotpink",
      "Baseline region" = "orange",
      "Other" = "grey70"
    ),
    name = "Region"
  ) +
  labs(
    x = "Calc_Ramp_Ex (nm)",
    y = "Defl_V_Ex (V)",
    title = "AFM Force Curve Regions"
  ) +
  theme_minimal(base_size = 14)
```


To analyze the baseline offset for each curve in an fdObj, we use the analyze_baseline() function. This function internally calls find_baseline() for each curve and records the computed baseline offset values in the metadata columns "baseline_V_approach" and "baseline_V_retract", along with the length of the baseline segment defined by the user.

Curves that fail baseline calculation are assigned NA in these metadata columns. Since baseline estimation requires a valid sensitivity value, any curve that fails sensitivity calculation will also fail baseline calculation. The portions of each raw curve used for baseline offset estimation are stored in the baseline_segment slot of the object.

```{r analyze baseline in batch}
test = analyze_baseline(test,useCurve = 'approach',least_length = 150)
test = analyze_baseline(test,useCurve = 'retract',least_length = 150)
view_df(test@metadata,'metadata with baseline values added as columns')
```


## Transformation of a raw curve

Now, with the sensitivity and baseline offset values calculated, we are ready to transform the curve with function transform_a_curve.

```{r single curve transform,fig.height=6,fig.width=16}
sc.1.tr = transform_a_curve(x = x, y= y, baseline = res.b$baseline, sensitivity = res$sensitivity,
                             spring_constant = 0.3188, senscal_seg_x = res$senscal_segment$x,
                             senscal_seg_y = res$senscal_segment$y)
p1 = ggplot(sc.1.raw, aes(x = Calc_Ramp_Ex_nm, y = Defl_V_Ex)) +
  geom_point(aes(
    color = case_when(
      `Region used for Sensitivity calculation` == TRUE ~ "Sensitivity region",
      `Region used for baseline offset calculation` == TRUE ~ "Baseline region",
      TRUE ~ "Other"
    )
  ), size = 1.5) +
  scale_color_manual(
    values = c(
      "Sensitivity region" = "hotpink",
      "Baseline region" = "orange",
      "Other" = "grey70"
    ),
    name = "Region"
  ) +
  labs(
    x = "Calc_Ramp_Ex (nm)",
    y = "Defl_V_Ex (V)",
    title = "AFM Force Curve Regions"
  ) +
  theme_minimal(base_size = 14) + 
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.5)

sc.1.tr$`Region used for Sensitivity calculation` = "FALSE"
sc.1.tr$`Region used for Sensitivity calculation`[1:nrow(res$senscal_segment)] <- 'TRUE'
sc.1.tr$`Region used for baseline offset calculation` = 'TRUE'
sc.1.tr$`Region used for baseline offset calculation`[1:(length(sc.1.raw$Defl_V_Ex)-150)] = 'FALSE'


p2 = ggplot(sc.1.tr, aes(x = separation_distance_nm, y = force_nN)) +
  geom_point(aes(
    color = case_when(
      `Region used for Sensitivity calculation` == TRUE ~ "Sensitivity region",
      `Region used for baseline offset calculation` == TRUE ~ "Baseline region",
      TRUE ~ "Other"
    )
  ), size = 1.5) +
  scale_color_manual(
    values = c(
      "Sensitivity region" = "hotpink",
      "Baseline region" = "orange",
      "Other" = "grey70"
    ),
    name = "Region"
  ) +
  labs(
    x = "Separation distance (nm)",
    y = "force (nN)",
    title = "AFM Force Curve Regions"
  ) +
  theme_minimal(base_size = 14) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.5)

plot_grid(p1,p2)
```

# Analytical metrics calculation

After curve transformation, we can calculate analytical metrics such as adhesive force, interaction energy and distance.

## Adhesive force

### Wrapper function for batch analysis

The adhesive force is calculated as the minimum force value in a curve, which represents the maximum attractive force between the AFM probe and the sample surface. The "analyze_curves_adhesive_force" function identifies this minimum force value and the separation distance at which it occurs, and records it in a new metadata column named "adhesive_force_nN_retract"/"adhesive_force_nN_approach" and another column named "adhesive_sep_nm_retract"/"adhesive_sep_nm_approach", depending on the selected curve portion via useCurve argument (either "approach" or "retract"). Curves that fail transformation will have NA in this column.

```{r calculate adhesive force}
test = analyze_curves_adhesive_force(test,useCurve = 'approach',threads = 1)
test = analyze_curves_adhesive_force(test,useCurve = 'retract',threads = 1)
view_df(test@metadata,'metadata with adhesive force values added as columns')   
```

### Single curve analysis
To demonstrate how the adhesive force is calculated for a single curve, we use the transformed retract curve of "180424_fc_mica_loc1.001" as an example. The function "analyze_a_curve_adhesive_force" identifies the minimum force value in the curve and the corresponding separation distance, and returns these values in a named vector.

```{r single curve adhesive force calculation}
sc.1.tr = test@retractCurves$`180424_fc_mica_loc1.001`
res.af = analyze_a_curve_adhesive_force(sc.1.tr)

print(res.af)
```

## Interaction energy

In AFM force spectroscopy, interaction energy is quantified by calculating the area under the curve. The analyze_a_curve_area() function computes adhesive energy (area below the negative noise threshold and above the curve inside Quadrant IV, representing bond-breaking energy) and repulsive energy (area above the positive noise threshold and under the curve inside Quadrant I, representing compression energy). A noise threshold is applied to filter out thermal fluctuations and instrumental noise, which would otherwise artificially inflate energy measurements. By excluding force values within the noise band, the function ensures that only biologically or physically meaningful interactions are counted, yielding more accurate energies.

### Wrapper function for batch analysis

The wrapper function "analyze_curves_energy" applies this analysis to all curves in an fdObj, and records the calculated adhesive energy and repulsive energy in new metadata columns named "adhesive_energy_fJ_retract"/"adhesive_energy_fJ_approach" and "repulsive_energy_fJ_retract"/"repulsive_energy_fJ_approach", depending on the selected curve portion via useCurve argument (either "approach" or "retract"). Curves that fail transformation will have NA in these columns.

```{r calculate interaction energy}
test = analyze_curves_energy(test,useCurve = 'approach',noise_cutoff = 0.5,threads = 1)
test = analyze_curves_energy(test,useCurve = 'retract',noise_cutoff = 0.5,threads = 1)
view_df(test@metadata,'metadata with energy values added as columns') 
```

### Single curve analysis

To demonstrate how the interaction energy is calculated for a single curve, we use the transformed retract curve of "180424_fv_pmon2_loc1-18-22.spm" as an example. The function "analyze_a_curve_energy" computes the adhesive energy and repulsive energy based on the area under the curve, applying a noise cutoff to exclude insignificant fluctuations.

```{r single curve interaction energy calculation}
pmon.1.tr = test@retractCurves$`180424_fv_pmon2_loc1-18-22.spm`
res.energy.noise_cutoff_at_0.5 = analyze_a_curve_area(pmon.1.tr,noise_cutoff = 0.5)
print(res.energy.noise_cutoff_at_0.5)

pmon.1.tr = test@retractCurves$`180424_fv_pmon2_loc1-18-22.spm`
res.energy.noise_cutoff_at_5 = analyze_a_curve_area(pmon.1.tr,noise_cutoff = 5)
print(res.energy.noise_cutoff_at_5)
```

```{r plot interaction energy with noise band, fig.height=6, fig.width=12}
p_0.5 <- plot_force_curve_energy(pmon.1.tr, noise_cutoff = 0.5)
p_5 <- plot_force_curve_energy(pmon.1.tr, noise_cutoff = 5)

plot_grid(p_0.5, p_5, ncol = 2)
```

## Interaction distance

The interaction distance is calculated as the separation distance at which the force curve first crosses a defined noise threshold, indicating the onset of significant interaction between the AFM probe and the sample surface. The "analyze_a_curve_interaction_distance" function identifies this point and returns the corresponding separation distance as the interaction distance.

### Wrapper function for batch analysis

The wrapper function "analyze_curves_interaction_distance" applies this analysis to all curves in an fdObj, and records the calculated interaction distance in a new metadata column named "interaction_distance_nm_retract"/"interaction_distance_nm_approach", depending on the selected curve portion via useCurve argument (either "approach" or "retract"). Curves that fail transformation will have NA in this column.

Here we set the noise threshold as 10 times of the standard deviation of the force values in the baseline region of each curve, which is a commonly used method to define the noise level in AFM force spectroscopy. Users can adjust the multiplier or use other methods via threshold_method argument to choose how the force cutoff is set from baseline noise—using scaled spread (sd, mad, iqr), directional quantiles (quantile), absolute-force quantile (abs_quantile), or a user-specified constant (fixed), with sd as the default.

In this call, baseline_span = "automatic" uses each curve’s stored baseline window from test@metadata$baseline_span_retract, y_direction = "negative" detects rupture-like (negative-force) excursions, and x_direction = "left" scans from the baseline side right-to-left to find the first threshold-crossing point.


```{r calculate interaction distance}
test = analyze_curves_interaction_distance(test,useCurve = 'retract',baseline_span = 'automatic',
                                           y_direction = "negative", x_direction = "left",
                                           multiplier = 10, threshold_method = "sd",
                                           threads = 1)

view_df(test@metadata,'metadata with interaction distance values added as columns')
```

similarly, we can calculate the interaction distance for the approach curves by changing the useCurve argument to "approach", and setting y_direction to "positive" to detect repulsive interaction

```{r calculate interaction distance for approach curves}
test = analyze_curves_interaction_distance(test,useCurve = 'approach',baseline_span = 'automatic',
                                           y_direction = "positive", x_direction = "right",
                                           multiplier = 10, threshold_method = "sd",
                                           threads = 1)
view_df(test@metadata,'metadata with interaction distance values for approach curves added as columns')
```


### Single curve analysis

To demonstrate how the interaction distance is calculated for a single curve, we use the transformed retract curve of "180424_fv_pmon2_loc1-18-22.spm" as an example. The function "analyze_a_curve_interaction_distance" identifies the point where the force curve first crosses the defined noise threshold and returns the corresponding separation distance as the interaction distance.


```{r single curve interaction distance calculation}
pmon.1.tr = test@retractCurves$`180424_fv_pmon2_loc1-18-22.spm`
res.interaction_distance = analyze_a_curve_interaction_distance(pmon.1.tr,baseline_span = 100, y_direction = "negative", x_direction = "left", multiplier = 10, threshold_method = "sd")

print(res.interaction_distance)
plot_force_curve_interaction_distance(pmon.1.tr, baseline_span = 100, y_direction = "negative", x_direction = "left",  multiplier = 10, threshold_method = "sd")
```


Similarly, we can calculate the interaction distance for the approach curve by changing the y_direction to "positive" and x_direction to "right" to detect repulsive interaction and scan from the baseline side left-to-right to find the first threshold-crossing point.

```{r single curve interaction distance calculation for approach curve}
pmon.1.approach = test@approachCurves$`180424_fv_pmon2_loc1-18-22.spm`
res.interaction_distance.approach = analyze_a_curve_interaction_distance(pmon.1.approach,baseline_span = 100, y_direction = "positive", x_direction = "right", multiplier = 10, threshold_method = "sd")

print(res.interaction_distance.approach)
plot_force_curve_interaction_distance(pmon.1.approach, baseline_span = 100, y_direction = "positive", x_direction = "right", multiplier = 10, threshold_method = "sd")

```


# Visualization of analytical metrics
```{r single curve plot metrics}
curve_name = '180424_fv_pmon2_loc1-18-22.spm'

plot_curve_metrics(
  fdobj = test,
  curve_name = curve_name,
  useCurve = 'retract',
  repulsive_energy = TRUE,
  adhesive_energy = TRUE,
  annotate_adhesive_force = TRUE,
  annotate_interaction_distance = TRUE,
  noise_cutoff = 0.5,
  interaction_type = 'auto',
  annotation_text_size = 5,
  interaction_label_color_rupture = 'purple4',
  interaction_label_color_repulsive = 'darkorange3',
  interaction_label_fill = 'white'
)
  ```

