---
title: "curvana: a comprehensive R package for analyzing atomic force microscopy force–distance curves"
subtitle: "Minimal workflow"
author: Yue (Gary) Zhang, PhD
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    number_sections: false
    highlight: tango
vignette: >
  %\VignetteIndexEntry{curvana tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=TRUE}
knitr::opts_chunk$set(warning = T, message = T,echo = T) 
```


# Introduction

This vignette shows a minimal workflow needed to analyze the raw curves. For information regarding the step-by-step demonstration of the full analytical workflow, please check out the full tutorial. 

# Preparation 

## Initializing work environment

```{r Initiate, results='hide',warning='False'}
packages_to_load = c('curvana','ggplot2','DT','dplyr','DT','ggpubr','cowplot')
invisible(suppressPackageStartupMessages({
    lapply(packages_to_load, library, character.only = TRUE)
}))
source(system.file("tutorial_helpers.R", package = "curvana"))
devtools::load_all()
```

# Load data

All examples below use the demo dataset provided in `extdata/` inside the package. To use your own dataset, you can pass the path to the directory holding the data to the "folder" argument. Note that users can increase the threads to enable parallel processing for loading a large folder (e.g., > 5000 curves), which will dramatically reduce the computational time.

At this step, we initialize the fdObj named "test" by importing the demo data directory "extdata/" using the "createFdObjFromFolder" function. An fdObj is the core data container in curvana, designed to store atomic force microscopy (AFM) force–distance curve data, including raw approach and retract curves, transformed data, associated metadata for downstream analysis, and the analytical metrics generated.

By default, createFdObjFromFolder will read all files in the designated folder with suffix ".txt", and expect each file has paired approach and retract deflection-dislocation raw curve data with standard column names (i.e., Calc_Ramp_Ex_nm, Defl_V_Ex, Calc_Ramp_Rt_nm, Defl_V_Rt). 


```{r load data}
#folder = system.file("extdata", package = "curvana")
#test.input = createFdObjFromFolder(folder = folder, threads = t)
input.path = "../../curvana_test/demo"
t = 1
test = createFdObjFromFolder(folder = input.path, threads = t)
```

Extract sample groups from the name of each raw curve file imported.

```{r edit metadata}
test@metadata <- test@metadata %>% 
  mutate(surface = sub("_.*","",filename),
         surface = case_when(
            surface == 'F' ~ 'fluorinated',
            surface == 'P' ~ 'PEG',
            surface == 'silicon' ~ 'silicon',
            surface == 'Z' ~ 'zwitterionic',
            TRUE ~ surface),
         region = sub(".*?_([0-9]+)\\.spm.*", "\\1", filename)
  )
view_df(test@metadata,'metadata')
```

# Preview the raw deflection curves

It is important to inspect the raw curves before formal analysis. We can use the "plot_deflection_curves" function to plot curves in batch. Here, we split curves into sub-panels by column "surface", and color code them by column "region".

```{r plot deflection curves, fig.height=12,fig.width=10}
plot_deflection_curves(test,split_curves_by = 'surface',group_curves_by = 'region',
                       alpha = 0.5,point_size = 0.5,
                       color_map = list(`1` = 'darkred',
                                        `2` = 'orange',
                                        `3` = 'steelblue'))
```


# Curve transformation

One advantage of curvana is that each step of the data analysis is modularized, enabling a flexible and customizable workflow. Curve transformation in curvana involves several steps, including sensitivity calculation, baseline offset calculation, and the application of these parameters to convert raw data into interaction force–separation distance curves. To execute all necessary steps at once, we can use a convenient function transform_curves.


Here we set the spring constant to 0.08 N/m for all curves, and set the least length of of the curve for baseline calculation to 200, which means the function will use the last 200 points of each curve to calculate the baseline offset. We also set the slope threshold to 0.01 and standard deviation threshold to 0.01 for filtering out curves with unstable baseline. For sensitivity calculation, as we have a small contact region, we set end = 20 to only use the first 20 points of the curve to calculate sensitivity.

```{r transform curves approach,warning=TRUE, message=TRUE,results='hold', echo=TRUE}
t = 1
# test_beforeTransform -> test
test = transform_curves(test,spring_constant = 0.08, # spring constant
                        useCurve = 'approach', # analyze the approach curve
                        # baseline determination and filtering
                        least_length = 200, # set the 200 data points at the far right side of each curve as the baseline region (the part of the curve presuming to be under zero force)
                        slp_threshold = 0.01, # any curves with the baseline region having slope > 0.01 will be discarded
                        std_threshold = 0.01, # any curves with the baseline region having standard deviation of y values > 0.01 will be discarded
                        # sensitivity calibration
                        end = 100, # use the first 100 data points at most on the left side to calibrate sensitivity
                        threads = t)

test = transform_curves(test,spring_constant = 0.08, # spring constant
                        useCurve = 'retract', # analyze the retract curve
                        # baseline determination and filtering
                        least_length = 200, # set the 200 data points at the far right side of each curve as the baseline region (the part of the curve presuming to be under zero force)
                        slp_threshold = 0.01, # any curves with the baseline region having slope > 0.01 will be discarded
                        std_threshold = 0.01, # any curves with the baseline region having standard deviation of y values > 0.01 will be discarded
                        # sensitivity calibration
                        end = 100, # use the first 100 data points at most on the left side to calibrate sensitivity
                        threads = t)
```

# Plot transformed curves

To visualize the transformed curves, we can use the "plot_fd_curves" function. Here we plot the approach and retract curves in separate panels, and color code them by column "surface".

```{r plot transformed curves, fig.height=12,fig.width=10}
plot_fd_curves(test,group_curves_by = 'region', split_curves_by = 'surface',
               point_alpha = 0.5,line_alpha = 0.3, point_size = 0.5,
               color_map = list(`1` = 'darkred',
                                `2` = 'orange',
                                `3` = 'steelblue'))
```

## Analytical metrics calculation

\code{analyze_curves_all_analytical_metrics()} is a wrapper function that performs all analytical metrics calculation at once. It first determines a baseline noise level so we can exclude insignificant fluctuations in metric calculations. After this, the adhesive force, adhesive and repulsive energy, and adhesive (rupture) and repulsive distance are determined.


\code{analyze_curves_all_analytical_metrics()}, and toggle adhesive-force or energy analysis on/off.

```{r run all analytical metrics with toggles}
test = analyze_curves_all_analytical_metrics(
  test, # the object
  useCurve = "both", # run analysis for both approach and retract 
  threads = 1, 
  noise_baseline_span = "automatic", # use the baseline section previously defined in curve transformation for noise band calculation
  noise_threshold_method = "quantile", # use quantile method to define the noise band
  noise_quantile_low = 0.05, # define the lower end of noise band as 5% percentile of y values of the baseline section
  noise_quantile_high = 1, # define the upper end of noise band as 95% percentile of y values of the baseline section
  analyze_adhesive_force = TRUE, # whether to analyze adhesive force
  analyze_energy = TRUE, # whether to analyze interaction energy (repulsive and adhesive)
  analyze_rupture_distance = TRUE, # whether to analyze adhesive/rupture distance
  analyze_repulsive_distance = TRUE, # whether to analyze repulsive distance
  analyze_rupture_distance_baseline_span = "automatic", # use the baseline section previously defined in curve transformation for rupture distance analysis
  analyze_rupture_distance_x_direction = "left", # scan from right to left to find the last data point before a curve enters the adhesive region from the noise band region
  analyze_repulsive_distance_baseline_span = "automatic", # use the baseline section previously defined in curve transformation for repulsive distance analysis
  analyze_repulsive_distance_x_direction = "right" # scan from left to right to find the last data point before a curve enters the noise band region from the repulsive region
)
```

# Summary statistics of analytical metrics

## Violin plot of selected metrics by surface

```{r summary statistics of metrics, fig.height=6, fig.width=12}
f1 = plot_metric_violin(test@metadata, metric_name = "adhesive_force_nN_retract", group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"))
f2 = plot_metric_violin(test@metadata, metric_name = "adhesive_sep_nm_retract", log10=T, group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"))
f3 = plot_metric_violin(test@metadata, metric_name = "adhesive_energy_aJ_retract",log10=T,  group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"))
f4 = plot_metric_violin(test@metadata, metric_name = "repulsive_energy_aJ_retract", group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"))
f5 = plot_metric_violin(test@metadata, metric_name = "rupture_distance_nm_retract", log10=T, group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"))
f6 = plot_metric_violin(test@metadata, metric_name = "repulsive_distance_nm_retract", log10=T, group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"))

plot_grid(f1,f2,f3,f4,f5,f6, nrow = 2)


```

## Violin plot with statistical annotations

```{r summary statistics with tests}
plot_metric_violin(
  df = test@metadata,
  metric_name = "adhesive_force_nN_retract",
  group_by = "surface",
  color_by = "surface",
  color_map = c(
    PEG = "#1b9e77",
    silicon = "#d95f02",
    fluorinated = "#7570b3",
    zwitterionic = "#e7298a"
  ),
  global_test = "anova",
  pairwise_test = "t.test",
  p_adjust_method = "BH"
)
```

## PCA biplot of selected metrics

```{r pca biplot of metrics, fig.height=8, fig.width=10}
pca_features <- c(
  "adhesive_force_nN_retract",
  "adhesive_sep_nm_retract",
  "adhesive_energy_aJ_retract",
  "repulsive_energy_aJ_retract",
  "rupture_distance_nm_retract",
  "repulsive_distance_nm_retract"
)

plot_pca_biplot(
  df = test@metadata,
  include_columns = pca_features,
  color_by = "surface",
  color_map = c(
    PEG = "#1b9e77",
    silicon = "#d95f02",
    fluorinated = "#7570b3",
    zwitterionic = "#e7298a"
  ),
  arrow_scale = 1.1,
  show_feature_labels = TRUE
)
```

## Scaled feature heatmap with sample annotations

The function `plot_complex_heatmap()` scales each selected feature by row (z-score),
plots features as rows and samples as columns, and can annotate columns by up to
two metadata variables.

```{r complex heatmap of metrics, fig.height=7, fig.width=10}
heatmap_features <- c(
  "adhesive_force_nN_retract",
  "adhesive_sep_nm_retract",
  "adhesive_energy_aJ_retract",
  "repulsive_energy_aJ_retract",
  "rupture_distance_nm_retract",
  "repulsive_distance_nm_retract"
)

  plot_complex_heatmap(
    df = test@metadata,
    include_columns = heatmap_features,
    annotate_columns = c("surface", "region"),
    annotation_colors = list(
      surface = c(
        PEG = "#1b9e77",
        silicon = "#d95f02",
        fluorinated = "#7570b3",
        zwitterionic = "#e7298a"
      )
    ),
    cluster_rows = TRUE,
    cluster_columns = FALSE,
    show_row_dend = TRUE,
    show_column_dend = FALSE,
    heatmap_name = "z-score",
    show_column_names = FALSE
  )
```

# Save all raw and transformed curves with analytical metrics annotated (optional)

```{r plot all metrics for a curve,fig.height=8, fig.width=16,eval=T}
plot_all_curve_metrics(
  fdobj = test, # the fdObj to plot
  useCurve = "retract", # plot the retract curve
  plot_raw = T, # whether to plot the raw curve
  destination_folder = '../test_plot_metrics', # the path to save the plots
  format = 'png', width = 18, height = 9, # plot format and dimensions
)
plot_all_curve_metrics(
  fdobj = test, # the fdObj to plot
  useCurve = "approach", # plot the approach curve
  plot_raw = T, # whether to plot the raw curve
  destination_folder = '../test_plot_metrics', # the path to save the plots
  format = 'png', width = 18, height = 9, # plot format and dimensions
)
```
