---
title: "curvana: a comprehensive R package for analyzing atomic force microscopy force–distance curves"
subtitle: "Minimal workflow"
author: Yue (Gary) Zhang, PhD
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    number_sections: false
    highlight: tango
vignette: >
  %\VignetteIndexEntry{curvana tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=TRUE}
knitr::opts_chunk$set(warning = T, message = T,echo = T) 
```


# Introduction

This vignette shows a minimal workflow needed to analyze the raw curves. For information regarding the step-by-step demonstration of the full analytical workflow, please check out the full tutorial. 

# Preparation 

## Initializing work environment

```{r Initiate, results='hide',warning='False'}
required_packages = c('curvana', 'ggplot2', 'DT', 'dplyr', 'cowplot')
optional_packages = c('ggpubr')

missing_required <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_required) > 0) {
  stop('Please install required packages first: ', paste(missing_required, collapse = ', '))
}

invisible(suppressPackageStartupMessages({
  lapply(required_packages, library, character.only = TRUE)
}))

available_optional <- optional_packages[vapply(optional_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(available_optional) > 0) {
  invisible(suppressPackageStartupMessages({
    lapply(available_optional, library, character.only = TRUE)
  }))
} else {
  message("Optional package 'ggpubr' is not installed; statistical annotation examples will be skipped.")
}

source(system.file("tutorial_helpers.R", package = "curvana"))
devtools::load_all("C:/Users/muson/coding/curvana")
```

# Load data

All examples below use the demo dataset provided in `extdata/` inside the package. To use your own dataset, you can pass the path to the directory holding the data to the "folder" argument. Note that users can increase the threads to enable parallel processing for loading a large folder (e.g., > 5000 curves), which will dramatically reduce the computational time.

At this step, we initialize the fdObj named "test" by importing the demo data directory "extdata/" using the "createFdObjFromFolder" function. An fdObj is the core data container in curvana, designed to store atomic force microscopy (AFM) force–distance curve data, including raw approach and retract curves, transformed data, associated metadata for downstream analysis, and the analytical metrics generated.

By default, createFdObjFromFolder will read all files in the designated folder with suffix ".txt", and expect each file has paired approach and retract deflection-dislocation raw curve data with standard column names (i.e., Calc_Ramp_Ex_nm, Defl_V_Ex, Calc_Ramp_Rt_nm, Defl_V_Rt). 


```{r load data}
folder = system.file("extdata", package = "curvana")
t = 1
test = createFdObjFromFolder(folder = folder, threads = t)

```

Extract sample groups from the name of each raw curve file imported.

```{r edit metadata}
test@metadata <- test@metadata %>% 
  mutate(surface = sub("_.*","",filename),
         surface = case_when(
            surface == 'F' ~ 'fluorinated',
            surface == 'P' ~ 'PEG',
            surface == 'silicon' ~ 'silicon',
            surface == 'Z' ~ 'zwitterionic',
            TRUE ~ surface),
         region = sub(".*?_([0-9]+)\\.spm.*", "\\1", filename)
  )
view_df(test@metadata,'metadata')
```

# Preview the raw deflection curves

It is important to inspect the raw curves before formal analysis. We can use the "plot_deflection_curves" function to plot curves in batch. Here, we split curves into sub-panels by column "surface", and color code them by column "region".

## Heatmap across each sampling event

Here we use the "plot_raw_deflection_heatmap" function to plot a heatmap of min-max scaled deflection values across each sampling event (i.e., each column is a curve, and each row is a data point along the curve). We also annotate columns by "surface" and "region", and set custom colors for each surface type.

This helps us visualize the overall pattern of deflection values across curves, and identify the baseline and sensitivity calibration regions for each curve. We can also check if there are any curves with abnormal patterns (e.g., unstable baseline, extremely high noise level) that may need to be excluded from downstream analysis.

```{r plot deflection heatmap, fig.height=12,fig.width=10,eval=TRUE}
plot_raw_deflection_heatmap(
  fdobj             = test,
  annotate_columns  = c("surface", "region"),
  annotation_colors = list(
    surface = c(PEG = "#1b9e77", silicon = "#d95f02",
                fluorinated = "#7570b3", zwitterionic = "#e7298a")
  )
)



```


## Preset baseline length

We can see that for all approach curves, deflection remains largely constant after approximately 100 data points. The total number data points per curve is 512, so we will set the last 400 points as baseline region for all curves. For retract curves, however, the fluorinated samples have a longer interaction region and shorter baseline, so we set their baseline region to the last 200 data points. For all other retract curves, we set the last 400 points to baseline regions.

```{r set baseline length}
test@metadata <- test@metadata %>%
  mutate(baseline_span_retract = case_when(
           surface == 'fluorinated' ~ 200,
           TRUE ~ 400
         ))
  

```

## Scatter plot of deflection curves

```{r plot deflection curves, fig.height=12,fig.width=10}
plot_deflection_curves(test,split_curves_by = 'surface',group_curves_by = 'region',
                       alpha = 0.5,point_size = 0.5,
                       color_map = list(`1` = 'darkred',
                                        `2` = 'orange',
                                        `3` = 'steelblue'))
```


One advantage of curvana is that each step of the data analysis is modularized, enabling a flexible and customizable workflow. Curve transformation in curvana involves several steps, including smoothing, sensitivity calculation, baseline offset calculation, and the application of these parameters to convert raw data into interaction force–separation distance curves. To execute all necessary steps at once, we can use a convenient function transform_curves.

Here we start with smoothing the raw curves with moving average of three consecutive points, set the spring constant to 0.08 N/m for all curves, and the slope threshold to 0.01 and standard deviation threshold to 0.01 for filtering out curves with unstable baseline. For sensitivity calculation, we set end = 100 to use at most the first 100 points of the curve to calculate sensitivity.

```{r transform curves approach,warning=TRUE, message=TRUE,results='hold', echo=TRUE}
t = 1
# test_beforeTransform -> test
test = transform_curves(test,
                        denoise_first = TRUE, # whether to apply denoising to raw curves before transformation. Denoising can help improve the reliability of sensitivity and baseline calculations, especially for noisy curves. The smoothed curve will be used for all downstream analysis, while the original raw curve will be preserved in the rawCurves slot of the object for reference.
                        p = 1, n = 3, # parameters for the Savitzky-Golay filter used in curve denoising. p is the polynomial order (default 1, which means linear), and n is the window size (default 3, which means each point will be smoothed by fitting a polynomial to itself and its 2 neighbors). Users can adjust these parameters based on the noise level and characteristics of their curves.
                        spring_constant = 0.08, # spring constant
                        useCurve = 'approach', # analyze the approach curve
                        # baseline determination and filtering
                        least_length = 400, # use the last 400 points as baseline
                        slp_threshold = 0.01, # any curves with the baseline region having slope > 0.01 will be discarded
                        std_threshold = 0.01, # any curves with the baseline region having standard deviation of y values > 0.01 will be discarded
                        # sensitivity calibration
                        intv = 4, # number of data points added per iteration when building the linear sensitivity segment
                        end = 60, # use the first 60 data points at most on the left side of the curveto calibrate sensitivity.
                        R_squared_min = 0.99, # minimum R^2 threshold for accepting a segment as sufficiently linear during sensitivity calculation
                        minimum_length = 4, # minimum number of accumulated points required for a valid sensitivity result; if shorter, sensitivity is reported as NA
                        threads = t)

test = transform_curves(test,
                        denoise_first = TRUE, # whether to apply denoising to raw curves before transformation. Denoising can help improve the reliability of sensitivity and baseline calculations, especially for noisy curves. The smoothed curve will be used for all downstream analysis, while the original raw curve will be preserved in the rawCurves slot of the object for reference.
                        p = 1, n = 3, # parameters for the Savitzky-Golay filter used in curve denoising. p is the polynomial order (default 1, which means linear), and n is the window size (default 3, which means each point will be smoothed by fitting a polynomial to itself and its 2 neighbors). Users can adjust these parameters based on the noise level and characteristics of their curves.
                        spring_constant = 0.08, # spring constant
                        useCurve = 'retract', # analyze the retract curve
                        # baseline determination and filtering
                        least_length = 'automatic', # set it to automatic so it will use numbers in baseline_span_retract column
                        slp_threshold = 0.01, # any curves with the baseline region having slope > 0.01 will be discarded
                        std_threshold = 0.01, # any curves with the baseline region having standard deviation of y values > 0.01 will be discarded
                        # sensitivity calibration
                        intv = 4, # number of data points added per iteration when building the linear sensitivity segment
                        end = 60, # use the first 60 data points at most on the left side of the curve to calibrate sensitivity.
                        R_squared_min = 0.99, # minimum R^2 threshold for accepting a segment as sufficiently linear during sensitivity calculation
                        minimum_length = 4, # minimum number of accumulated points required for a valid sensitivity result; if shorter, sensitivity is reported as NA
                        threads = t)
```

# Plot transformed curves

To visualize the transformed curves, we can use the "plot_fd_curves" function. Here we plot the approach and retract curves in separate panels, and color code them by column "surface".

```{r plot transformed curves, fig.height=12,fig.width=10}
plot_fd_curves(test,group_curves_by = 'region', split_curves_by = 'surface',
               point_alpha = 0.5,line_alpha = 0.3, point_size = 0.5,
               color_map = list(`1` = 'darkred',
                                `2` = 'orange',
                                `3` = 'steelblue'))
```

# Analytical metrics calculation

\code{analyze_curves_all_analytical_metrics()} is a wrapper function that performs all analytical metrics calculation at once. It first determines a baseline noise level so we can exclude insignificant fluctuations in metric calculations. After this, the adhesive force, adhesive and repulsive energy, and adhesive (rupture) and repulsive distance are determined.


\code{analyze_curves_all_analytical_metrics()}, and toggle adhesive-force or energy analysis on/off.

```{r run all analytical metrics with toggles}
test = analyze_curves_all_analytical_metrics(
  test, # the object
  useCurve = "both", # run analysis for both approach and retract 
  threads = 1, 
  noise_baseline_span = "automatic", # use the baseline section previously defined in curve transformation for noise band calculation
  noise_threshold_method = "quantile", # use quantile method to define the noise band
  noise_quantile_low = 0.00, # define the lower end of noise band as 0% percentile of y values of the baseline section
  noise_quantile_high = 1, # define the upper end of noise band as 100% percentile of y values of the baseline section
  noise_multiplier = 1, # multiply the noise band by 2 to get a more stringent threshold for metric analysis
  analyze_adhesive_force = TRUE, # whether to analyze adhesive force
  analyze_energy = TRUE, # whether to analyze interaction energy (repulsive and adhesive)
  analyze_rupture_distance = TRUE, # whether to analyze adhesive/rupture distance
  analyze_repulsive_distance = TRUE, # whether to analyze repulsive distance
  analyze_rupture_distance_baseline_span = "automatic", # use the baseline section previously defined in curve transformation for rupture distance analysis
  analyze_rupture_distance_x_direction = "left", # scan from right to left to find the first data point at which a curve enters the adhesive region from the noise band region
  analyze_rupture_distance_min_consecutive = 3, # to reduce false positives from random fluctuations, require at least 3 consecutive points below the lower noise-band threshold for classifying the curve as entering the adhesive region.
  analyze_repulsive_distance_baseline_span = "automatic", # use the baseline section previously defined in curve transformation for repulsive distance analysis
  analyze_repulsive_distance_x_direction = "right" # scan from left to right to find the last data point before a curve enters the noise band region from the repulsive region
)
```

All calculated analytical metrics will be stored in the metadata table of the fdObj, and can be accessed by "test@metadata". The names of columns for each metric are self-explanatory, and include the metric name, the unit, and the curve type used for analysis (approach or retract).

```{r view metrics}
view_df(test@metadata,'metadata')
```

# View selected curves with metrics annotated

To visualize the curves with analytical metrics annotated, we can use the "plot_curve_with_metrics" function. Here we plot the retract curve of a selected sample, with setting plot_raw = TRUE to display the raw data side by side with the annotated metrics.

```{r plot curve with metrics, fig.height=6, fig.width=15}
plot_a_curve_metrics(
  fdobj = test,
  curve_name = 'F_1.spm-F4E1_ForceCurveIndex_1.spm'	,
  useCurve = "retract",plot_raw = T,base_size = 15,annotation_text_size = 4, point_alpha = 0.5
)
```


# Summary statistics of analytical metrics

## PCA biplot of selected metrics

As we have multiple analytical metrics, we can use PCA to visualize the overall pattern of how samples cluster based on these metrics, and which metrics contribute the most to the clustering pattern. The function "plot_pca_biplot" performs PCA analysis and plots the PCA biplot. Here we color code samples by column "surface", and set custom colors for each surface type. We also set "arrow_scale" to 1.1 to make the feature arrows longer for better visualization, and set "show_feature_labels" to TRUE to show the names of features on the plot.

```{r pca biplot of metrics, fig.height=8, fig.width=10}
pca_features <- c(
  "adhesive_force_nN_retract",
  "adhesive_energy_aJ_retract",
  "repulsive_energy_aJ_retract",
  "rupture_distance_nm_retract",
  "repulsive_distance_nm_retract"
)

plot_pca_biplot(
  df = test@metadata,
  include_columns = pca_features,
  color_by = "surface",
  color_map = c(
    PEG = "#1b9e77",
    silicon = "#d95f02",
    fluorinated = "#7570b3",
    zwitterionic = "#e7298a"
  ),
  arrow_scale = 1.1,
  show_feature_labels = TRUE
)
```


## Violin plot of selected metrics by surface

To visualize the distribution of each metric across different sample groups, we can use the "plot_metric_violin" function to plot violin plots. Here we plot the distribution of adhesive force, the separation distance at which the maximum adhesive force occurs, adhesive energy, repulsive energy, rupture distance, and repulsive distance for each surface type. We also set log10 = TRUE for metrics with a wide range of values to improve visualization.

```{r summary statistics of metrics, fig.height=5, fig.width=18}
f1 = plot_metric_violin(test@metadata, metric_name = "adhesive_force_nN_retract", group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"))
f2 = plot_metric_violin(test@metadata, metric_name = "adhesive_energy_aJ_retract",log10=F,  group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"))
f3 = plot_metric_violin(test@metadata, metric_name = "repulsive_energy_aJ_retract", group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"))
f4 = plot_metric_violin(test@metadata, metric_name = "rupture_distance_nm_retract", log10=T, group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"))
f5 = plot_metric_violin(test@metadata, metric_name = "repulsive_distance_nm_retract", log10=T, group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"))

plot_grid(f1,f2,f3,f4,f5, nrow = 1)


```

## Violin plot with statistical annotations

We can also add statistical annotations to the violin plots to show the significance of differences between groups. Here we perform ANOVA as the global test to determine if there are overall differences between groups, and then perform pairwise t-tests with Benjamini-Hochberg adjustment for multiple comparisons to determine which specific groups are different from each other. The significant levels will be annotated on the plot.

```{r summary statistics with tests, eval=requireNamespace("ggpubr", quietly = TRUE)}
plot_metric_violin(
  df = test@metadata,
  metric_name = "adhesive_force_nN_retract",
  group_by = "surface",
  color_by = "surface",add_points = F,
  show_whisker_box = T,
  color_map = c(
    PEG = "#1b9e77",
    silicon = "#d95f02",
    fluorinated = "#7570b3",
    zwitterionic = "#e7298a"
  ),
  global_test = "anova",
  pairwise_test = "t.test",
  p_adjust_method = "BH"
)
```


## Scaled feature heatmap with sample annotations

The function `plot_complex_heatmap()` scales each selected feature by row (z-score),
plots features as rows and samples as columns, and can annotate columns by up to
two metadata variables.

```{r complex heatmap of metrics, fig.height=7, fig.width=10}
heatmap_features <- c(
  "adhesive_force_nN_retract",
  "adhesive_energy_aJ_retract",
  "repulsive_energy_aJ_retract",
  "rupture_distance_nm_retract",
  "repulsive_distance_nm_retract"
)

  plot_complex_heatmap(
    df = test@metadata,
    include_columns = heatmap_features,
    annotate_columns = c("surface", "region"),
    annotation_colors = list(
      surface = c(
        PEG = "#1b9e77",
        silicon = "#d95f02",
        fluorinated = "#7570b3",
        zwitterionic = "#e7298a"
      )
    ),
    cluster_rows = TRUE,
    cluster_columns = FALSE,
    show_row_dend = TRUE,
    show_column_dend = FALSE,
    heatmap_name = "z-score",
    show_column_names = FALSE
  )
```

# Save all raw and transformed curves with analytical metrics annotated (optional)

```{r plot all metrics for a curve,fig.height=8, fig.width=16,eval=T}
plot_all_curve_metrics(
  fdobj = test, # the fdObj to plot
  useCurve = "retract", # plot the retract curve
  plot_raw = T, # whether to plot the raw curve
  destination_folder = '../test_plot_metrics_denoised', # the path to save the plots
  format = 'png', width = 18, height = 9, # plot format and dimensions,
  threads = 8 # number of threads to use for parallel processing (set to 4 to use 4 cores, which will significantly speed up the plotting process for a large dataset
)
plot_all_curve_metrics(
  fdobj = test, # the fdObj to plot
  useCurve = "approach", # plot the approach curve
  plot_raw = T, # whether to plot the raw curve
  destination_folder = '../test_plot_metrics_denoised', # the path to save the plots
  format = 'png', width = 18, height = 9, # plot format and dimensions
  threads = 8 # number of threads to use for parallel processing (set to 4 to use 4 cores, which will significantly speed up the plotting process for a large dataset
)
```
