---
title: "curvana: a comprehensive R package for analyzing atomic force microscopy force–distance curves"
subtitle: "Minimal workflow"
author: Yue (Gary) Zhang, PhD
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    number_sections: false
    highlight: tango
vignette: >
  %\VignetteIndexEntry{curvana tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=TRUE}
knitr::opts_chunk$set(warning = T, message = T,echo = T) 
```


# Introduction

This vignette shows a minimal workflow needed to analyze the data from loading raw curves to generating summary statistics and visualizations. For information regarding the step-by-step demonstration of the full analytical workflow, please check out `vignette_full_workflow.html`. 

# Preparation 

## Initializing work environment

```{r Initiate, results='hide',warning='False'}
required_packages = c('curvana', 'ggplot2', 'DT', 'dplyr', 'cowplot')
optional_packages = c('ggpubr')

missing_required <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_required) > 0) {
  stop('Please install required packages first: ', paste(missing_required, collapse = ', '))
}

invisible(suppressPackageStartupMessages({
  lapply(required_packages, library, character.only = TRUE)
}))

available_optional <- optional_packages[vapply(optional_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(available_optional) > 0) {
  invisible(suppressPackageStartupMessages({
    lapply(available_optional, library, character.only = TRUE)
  }))
} else {
  message("Optional package 'ggpubr' is not installed; statistical annotation examples will be skipped.")
}

source(system.file("tutorial_helpers.R", package = "curvana"))
devtools::load_all() 
```

# Load data


## Example of raw curve data file

Here we provide an example of an input file. The file contains two columns: "Calc_Ramp_Ex_nm" and "Defl_V_Ex", which are the piezo displacement in nm and cantilever deflection in V for the approach segment of the curve, respectively. The file also has a column named "Calc_Ramp_Rt_nm" and "Defl_V_Rt" for the retract segment. 

Note that the direction of the curve is the contact region at the end (right side) and the non-interaction region at the beginning (left side), which is the default format of Bruker AFM output. However, by convention, the contact region is usually at the beginning of the curve and the non-interaction region is at the end of the curve. To correct this, we can set reverse to True for the corresponding columns (see the next section).

```{r example raw curve}
curve_path <- system.file("extdata", "F_1.spm-F4E1_ForceCurveIndex_0.spm.txt", package = "curvana")
curve_df <- read.table(curve_path, header = TRUE, sep = "\t")
view_df(curve_df,'Example raw curve data')

print(ggplot(curve_df) +
  geom_line(aes(x = Calc_Ramp_Ex_nm, y = Defl_V_Ex, color = "Approach")) +
  geom_line(aes(x = Calc_Ramp_Rt_nm, y = Defl_V_Rt, color = "Retract")) +
  scale_color_manual(values = c("Approach" = "steelblue", "Retract" = "tomato")) +
  labs(x = "Piezo Displacement (nm)", y = "Deflection (V)", color = "Segment") +
  theme_classic())

```


## Load demo data

All examples below use the demo dataset provided in `extdata/` inside the package. To use your own dataset, you can pass the path to the directory holding the data to the "folder" argument. Note that users can increase the threads to enable parallel processing for loading a large folder (e.g., > 5000 curves), which will substantial reduce the computational time.

At this step, we initialize the fdObj named "test" by importing the demo data directory "extdata/" using the "createFdObjFromFolder" function. An fdObj is the core data container in curvana, designed to store atomic force microscopy (AFM) force–distance curve data, including raw curves, transformed data, associated metadata for downstream analysis, and the analytical metrics generated.

By default, createFdObjFromFolder will read all files in the designated folder with suffix ".txt", and expect each file has at least one pair of piezo displacement and cantilever deflection values for approach or retract segments. The name of the columns can be specified by the user, but by default the function will look for columns with standard column names (i.e., Calc_Ramp_Ex_nm, Defl_V_Ex, Calc_Ramp_Rt_nm, Defl_V_Rt) output by Bruker AFM. Users working with non-Bruker input files can specify column headers. Each curve should have the contact region at the beginning and the non-interaction region at the end. If any column needs to be reversed, you can set reverse to True for that column.

If users only have either approach or retract data, they can also use this function to create an fdObj by only specifying the column names for the available segment, and leave the argument related to the missing segment as is. The function will automatically disregard the missing segment and create an fdObj with only the available segment.

```{r load data}
folder = system.file("extdata", package = "curvana")
t = 1
test.input = createFdObjFromFolder(folder = folder, threads = t,
                                  Displacement_Approach = "Calc_Ramp_Ex_nm", # column name for piezo displacement in approach curve
                                  Deflection_Approach = "Defl_V_Ex", # column name for cantilever deflection in approach curve
                                  Displacement_Retract = "Calc_Ramp_Rt_nm", # column name for piezo displacement in retract curve
                                  Deflection_Retract = "Defl_V_Rt", # column name for cantilever deflection in retract curve
                                  reverse_Displacement_Approach = FALSE, # set to FALSE to not reverse the column for piezo displacement in approach curve
                                  reverse_Displacement_Retract = TRUE, # set to TRUE to reverse the column for piezo displacement in retract curve
                                  reverse_Deflection_Approach = TRUE, # set to TRUE to reverse the column for cantilever deflection in approach curve
                                  reverse_Deflection_Retract = FALSE # set to FALSE to not reverse the column for cantilever deflection in retract curve
                                  )

```

Let's take a quick look at the fdObj (named test) we created. We can see that we imported 10 raw curves. Non of them are transformed yet, and "filename" is the only column in the metadata slot.

```{r show init fdobj}
print(test.input)
view_df(test.input@metadata,'metadata')
```


Extract sample groups from the name of each raw curve file imported.

```{r edit metadata}
test <- test.input
test@metadata <- test@metadata %>% 
  mutate(surface = sub("_.*","",filename),
         surface = case_when(
            surface == 'F' ~ 'fluorinated',
            surface == 'P' ~ 'PEG',
            surface == 'silicon' ~ 'silicon',
            surface == 'Z' ~ 'zwitterionic',
            TRUE ~ surface),
         region = sub(".*?_([0-9]+)\\.spm.*", "\\1", filename)
  )
view_df(test@metadata,'metadata')
test_beforeTransform <- test
```


# Preview the raw deflection curves

It is important to inspect the raw curves before formal analysis. Here we visualize the raw curves with a heatmap and a scatter plot to check the overall pattern of the curves.

## Heatmap across each sampling event

Here we use the "plot_raw_deflection_heatmap" function to plot a heatmap of min-max scaled deflection values across each sampling event (i.e., each column is a curve, and each row is a data point along the curve). We also annotate columns by "surface" and "region", and set custom colors for each surface type.

This helps us visualize the overall pattern of deflection values across curves, and identify the baseline and sensitivity calibration regions for each curve. 

```{r plot deflection heatmap, fig.height=6,fig.width=10,eval=TRUE}
plot_raw_deflection_heatmap(
  fdobj             = test,
  annotate_columns  = c("surface", "region"),
  annotation_colors = list(
    surface = c(PEG = "#1b9e77", silicon = "#d95f02",
                fluorinated = "#7570b3", zwitterionic = "#e7298a")
  )
)



```


## Scatter plot of deflection curves

We can use the "plot_deflection_curves" function to plot curves in batch. Here, we split curves into sub-panels by column "surface", and color code them by column "region".

```{r plot deflection curves, fig.height=8,fig.width=10}
plot_deflection_curves(test,split_curves_by = 'surface',group_curves_by = 'region',
                       alpha = 0.5,point_size = 0.5,
                       color_map = list(`1` = 'darkred',
                                        `2` = 'orange',
                                        `3` = 'steelblue'))
```

# Curve transformation

Curve transformation in curvana involves several steps, including smoothing, sensitivity calculation, baseline offset calculation, and the application of these parameters to convert raw data into interaction force–separation distance curves. To execute all necessary steps at once, we can use a convenient function transform_curves.

```{r transform curves approach,warning=TRUE, message=TRUE,results='hold', echo=TRUE}
# transformation for the approach curves
test = transform_curves(test,
                        useCurve = 'approach', # analyze the approach curve

                        denoise_first = TRUE, # whether to apply denoising to raw curves before transformation. Denoising can help improve the reliability of sensitivity and baseline calculations, especially for noisy curves. The smoothed curve will be used for all downstream analysis, while the original raw curve will be preserved in the rawCurves slot of the object for reference.
                        p = 1, n = 3, # parameters for the Savitzky-Golay filter used in curve denoising. p is the polynomial order (default 1, which means linear), and n is the window size (default 3, which means each point will be smoothed by fitting a polynomial to itself and its 2 neighbors). Users can adjust these parameters based on the noise level and characteristics of their curves. This setting is equivalent to a simple moving average.
                        
                        spring_constant = 0.08, # spring constant; can be a single numeric value applied to all curves, or the name of a metadata column containing per-curve spring constant values.

                        # baseline determination and filtering
                        least_length = 300, # use the last 300 points as baseline
                        slp_threshold = 0.01, # any curves with the baseline region having slope > 0.01 will be discarded
                        std_threshold = 0.01, # any curves with the baseline region having standard deviation of y values > 0.01 will be discarded

                        # sensitivity calibration
                        intv = 4, # number of data points added per iteration when building the linear sensitivity segment
                        end = 20, # use the first 20 data points at most on the left side of the curve to calibrate sensitivity.
                        R_squared_min = 0.99, # minimum R^2 threshold for accepting a segment as sufficiently linear during sensitivity calculation
                        minimum_length = 4, # minimum number of accumulated points required for a valid sensitivity result; if shorter, sensitivity is reported as NA
                        threads = 1 # number of threads to use for parallel processing; set to 1 for no parallelization; increase for faster processing of large datasets (e.g., > 5000 curves)
                        )

# repeat the same transformation for the retract curves
test = transform_curves(test,
                        useCurve = 'retract', # analyze the retract curve
                        denoise_first = TRUE, # whether to apply denoising to raw curves before transformation. Denoising can help improve the reliability of sensitivity and baseline calculations, especially for noisy curves. The smoothed curve will be used for all downstream analysis, while the original raw curve will be preserved in the rawCurves slot of the object for reference.
                        p = 1, n = 3, # parameters for the Savitzky-Golay filter used in curve denoising. p is the polynomial order (default 1, which means linear), and n is the window size (default 3, which means each point will be smoothed by fitting a polynomial to itself and its 2 neighbors). Users can adjust these parameters based on the noise level and characteristics of their curves. This setting is equivalent to a simple moving average.
                        
                        spring_constant = 0.08, # spring constant; can be a single numeric value applied to all curves, or the name of a metadata column containing per-curve spring constant values.
                        
                        # baseline determination and filtering
                        least_length = 300, # use the last 300 points as baseline
                        slp_threshold = 0.01, # any curves with the baseline region having slope > 0.01 will be discarded
                        std_threshold = 0.01, # any curves with the baseline region having standard deviation of y values > 0.01 will be discarded

                        # sensitivity calibration
                        intv = 4, # number of data points added per iteration when building the linear sensitivity segment
                        end = 20, # use the first 20 data points at most on the left side of the curve to calibrate sensitivity.
                        R_squared_min = 0.99, # minimum R^2 threshold for accepting a segment as sufficiently linear during sensitivity calculation
                        minimum_length = 4, # minimum number of accumulated points required for a valid sensitivity result; if shorter, sensitivity is reported as NA
                        threads = 1 # number of threads to use for parallel processing; set to 1 for no parallelization; increase for faster processing of large datasets (e.g., > 5000 curves)
                        )
```

## Soft surface correction

__When measuring on a compliant (soft) surface — such as a hydrogel, cell membrane, or polymer brush — the contact region of the curve reflects a compound response of both cantilever deflection and surface indentation. Using this region to estimate probe sensitivity will therefore underestimate the true probe sensitivity. To correct for this, the probe sensitivity should be calibrated separately on a rigid reference surface (e.g., bare glass or silicon) and then supplied to transform_curves() when processing the soft-surface curves.__

__To do this, set soft = TRUE and provide the externally calibrated sensitivity via probe_sensitivity_external. This can be either a single numeric value (applied to all curves) or the name of a metadata column containing per-curve sensitivity values. When soft = TRUE, transform_curves() uses this external sensitivity to convert deflection (V) to force (nN). Note that the x-axis transformation (piezo displacement to separation distance) is unaffected and always uses the sensitivity estimated from each individual curve, preserving the deformation component in the separation distance.__



```{r use external sensitivity value}

test3 = transform_curves(test,
                        useCurve = 'retract', # analyze the retract curve
                        denoise_first = TRUE, 
                        p = 1, n = 3, 
                        spring_constant = 0.08, 
                        # baseline determination and filtering
                        least_length = 300, 
                        std_threshold = 0.01, 
                        slp_threshold = 0.01,
                        # sensitivity calibration
                        intv = 4, 
                        end = 20, 
                        R_squared_min = 0.99, 
                        minimum_length = 4, 
                        threads = 1,

                        # soft surface correction
                        soft = TRUE, # whether to apply soft surface correction
                        probe_sensitivity_external = 0.05 # externally calibrated sensitivity value to use for soft surface correction (in V/nm). This can be a single numeric value applied to all curves, or the name of a metadata column containing per-curve sensitivity values.
                        )
```

# Plot transformed curves

To visualize the transformed curves, we can use the "plot_fd_curves" function. Here we plot the approach and retract curves in separate panels, and color code them by column "surface".

```{r plot transformed curves, fig.height=12,fig.width=10}
plot_fd_curves(test,group_curves_by = 'region', split_curves_by = 'surface',
               point_alpha = 0.5,line_alpha = 0.3, point_size = 0.5,
               color_map = list(`1` = 'darkred',
                                `2` = 'orange',
                                `3` = 'steelblue'))
```

We can see that Fluorinated surface has the largest adhesion, which occurs mostly immediately after pulling probe off the surface. The other surfaces also have adhesion, but the magnitude of adhesion is smaller. All curves passed the transformation step have a baseline that is well aligned with the zero-force line with a relatively small fluctuation, which is important for accurate calculation of analytical metrics in later steps.

# Analytical metrics calculation

`analyze_curves_all_analytical_metrics()` is a wrapper function that performs all analytical metrics calculation at once. It first determines a baseline noise level so we can exclude insignificant fluctuations in metric calculations. After this, the adhesive force, adhesive and repulsive energy, and adhesive (rupture) and repulsive distance are determined.


```{r run all analytical metrics with toggles}
test = analyze_curves_all_analytical_metrics(
  test, # the object
  useCurve = "both", # run analysis for both approach and retract 
  threads = 1, 

  # noise band parameters
  noise_baseline_span = "automatic", # use the baseline section previously defined in curve transformation for noise band calculation
  noise_threshold_method = "quantile", # use quantile method to define the noise band
  noise_quantile_low = 0.00, # define the lower end of noise band as minimum y value of the baseline section
  noise_quantile_high = 1, # define the upper end of noise band as maximum y value of the baseline section
  noise_multiplier = 1, # multiply the noise band by 1 to directly use the min-max range of the force in baseline region without further scaling.

  # adhesive force
  analyze_adhesive_force = TRUE, # whether to analyze adhesive force
  # adhesive energy and repulsive energy
  analyze_energy = TRUE, # whether to analyze interaction energy (repulsive and adhesive)
  # rupture (adhesive) distance
  analyze_rupture_distance = TRUE, # whether to analyze adhesive/rupture distance
  # repulsive distance
  analyze_repulsive_distance = TRUE # whether to analyze repulsive distance
)
```

All calculated analytical metrics will be stored in the metadata table of the fdObj, and can be accessed by `test@metadata`. The names of columns for each metric are self-explanatory, and include the metric name, the unit, and the curve type used for analysis (approach or retract).

```{r view metrics}
view_df(test@metadata,'metadata')
```

# View selected curves with metrics annotated

To visualize the curves with analytical metrics annotated, we can use the "plot_curve_with_metrics" function. Here we plot the retract curve of a selected sample, with setting plot_raw = TRUE to display the raw data side by side with the annotated metrics.

```{r plot curve with metrics, fig.height=12, fig.width=22}
plot_a_curve_metrics(
  fdobj = test,
  curve_name = 'F_1.spm-F4E1_ForceCurveIndex_1.spm'	,
  useCurve = "retract", # which curve to plot, can be "approach" or "retract".
  plot_raw = T, # whether to plot the raw curve together with the transformed curve. 
  base_size = 15,
  annotation_text_size = 5
)
```

# Save all raw and transformed curves with analytical metrics annotated (optional)

```{r plot all metrics for a curve,fig.height=8, fig.width=16,eval=F}
plot_all_curve_metrics(
  fdobj = test, # the fdObj to plot
  useCurve = "retract", # plot the retract curve
  plot_raw = T, # whether to plot the raw curve
  destination_folder = '../test_plot_metrics_denoised', # the path to save the plots
  format = 'png', width = 18, height = 9, # plot format and dimensions,
  threads = 8 # number of threads to use for parallel processing (set to 8 to use 8 cores, which will significantly speed up the plotting process for a large dataset
)
plot_all_curve_metrics(
  fdobj = test, # the fdObj to plot
  useCurve = "approach", # plot the approach curve
  plot_raw = T, # whether to plot the raw curve
  destination_folder = '../test_plot_metrics_denoised', # the path to save the plots
  format = 'png', width = 18, height = 9, # plot format and dimensions
  threads = 8 # number of threads to use for parallel processing (set to 8 to use 8 cores, which will significantly speed up the plotting process for a large dataset
)
```

# Summary statistics of analytical metrics

## PCA biplot of selected metrics

Dimensional reduction is first used to visualize the overall pattern of how curves cluster based on their analytical metrics. The function "plot_pca_biplot" performs principal component analysis (PCA) on the selected metrics and plots the scores of samples in the PCA space, colored by a specified metadata variable (e.g., surface type). The function also plots the loadings of the selected features as arrows, indicating how each metric contributes to the principal components and the clustering pattern of samples.

```{r pca biplot of metrics, fig.height=6, fig.width=7}
pca_features <- c(
  "adhesive_force_nN_retract",
  "adhesive_energy_aJ_retract",
  "rupture_distance_nm_retract",
  "repulsive_energy_aJ_approach",
  "repulsive_distance_nm_retract"
)

plot_pca_biplot(
  df = test@metadata,
  include_columns = pca_features,
  color_by = "surface",
  color_map = c(
    PEG = "#1b9e77",
    silicon = "#d95f02",
    fluorinated = "#7570b3",
    zwitterionic = "#e7298a"
  ),
  arrow_scale = 1.1,
  show_feature_labels = TRUE
)
```

From the PCA biplot, we can see the fluorinated surface is well separated from the other three surfaces, which is consistent with our observation of the curves and the fact that fluorinated surface has the largest adhesion. The biplot also shows that adhesive force, adhesive energy and rupture distance are the main contributors to PC1, which separates fluorinated surface from the others, while repulsive energy and repulsive distance contribute more to PC2, which largely separates zwitterionic surface from PEG and silicon surfaces. Adhesive force, adhesive energy and rupture distance are positively correlated with each other, which is consistent with the fact that they all reflect the strength of adhesion. Repulsive energy and repulsive distance are also positively correlated, which makes sense because a stronger repulsion would typically extend over a longer distance.


## Scaled feature heatmap with sample annotations

The function `plot_complex_heatmap()` creates a heatmap of selected analytical metrics across samples, with options for clustering and annotation. In this example, we select the same metrics used for PCA plot to visualize their relative values across samples. The heatmap is scaled by z-score to show how each sample deviates from the mean for each metric, allowing for easy comparison across different metrics with different units and ranges. The samples are annotated by surface type and region, with a specified color map for the surface types. Clustering is applied to rows (samples) to reveal patterns of similarity based on the selected metrics.

```{r complex heatmap of metrics, fig.height=5, fig.width=8}
  plot_complex_heatmap(
    df = test@metadata,
    include_columns = pca_features,
    annotate_columns = c("surface", "region"),
    annotation_colors = list(
      surface = c(
        PEG = "#1b9e77",
        silicon = "#d95f02",
        fluorinated = "#7570b3",
        zwitterionic = "#e7298a"
      )
    ),
    cluster_rows = TRUE,
    cluster_columns = FALSE,
    show_row_dend = TRUE,
    show_column_dend = FALSE,
    heatmap_name = "z-score",
    show_column_names = FALSE
  )
```

## Violin plot of selected metrics by surface

`plot_metric_violin()` can be used to visualize the distribution of selected metrics across different surfaces. Here we plot the violin plot of adhesive force, adhesive energy and rupture distance of the retract curves, as well as the repulsive energy and repulsive distance of approach curves. The option "show_whisker_box" is set to TRUE to show the median and interquartile range of the distribution.

```{r summary statistics of metrics, fig.height=6, fig.width=12}
f1 = plot_metric_violin(test@metadata, metric_name = "adhesive_force_nN_retract", group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"),
show_whisker_box = T)
f2 = plot_metric_violin(test@metadata, metric_name = "adhesive_energy_aJ_retract",log10=T,  group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"),
show_whisker_box = T)
f3 = plot_metric_violin(test@metadata, metric_name = "repulsive_energy_aJ_approach", group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"),
show_whisker_box = T)
f4 = plot_metric_violin(test@metadata, metric_name = "rupture_distance_nm_retract", log10=T, group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"),
show_whisker_box = T)
f5 = plot_metric_violin(test@metadata, metric_name = "repulsive_distance_nm_approach", log10=T, group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"),
show_whisker_box = T)

plot_grid(f1,f2,f3,f4,f5, nrow = 2)


```

## Violin plot with statistical annotations

`plot_metric_violin()` also has options to perform statistical tests and annotate the significance on the plot. Here we perform Kruskal-Wallis test to compare the means of adhesive force across different surfaces, and perform pairwise Wilcoxon tests for post-hoc analysis. The p-values are adjusted using Benjamini-Hochberg method to control for false discovery rate.

```{r summary statistics with tests, eval=requireNamespace("ggpubr", quietly = TRUE)}
plot_metric_violin(
  df = test@metadata,
  metric_name = "adhesive_force_nN_retract",
  group_by = "surface",
  color_by = "surface",
  color_map = c(
    PEG = "#1b9e77",
    silicon = "#d95f02",
    fluorinated = "#7570b3",
    zwitterionic = "#e7298a"
  ),
  global_test = "kruskal",
  pairwise_test = "wilcox",
  p_adjust_method = "BH"

)
```

The function also allows users to specify which pairwise comparisons to perform, so that only relevant comparisons are made and the plot is not overcrowded with too many significance annotations. Here we specify to compare fluorinated surface with each of the other three surfaces.

```{r summary statistics with tests2, eval=requireNamespace("ggpubr", quietly = TRUE)}
plot_metric_violin(
  df = test@metadata,
  metric_name = "adhesive_force_nN_retract",
  group_by = "surface",
  color_by = "surface",
  color_map = c(
    PEG = "#1b9e77",
    silicon = "#d95f02",
    fluorinated = "#7570b3",
    zwitterionic = "#e7298a"
  ),
  global_test = "none", # do not perform global test, only do pairwise comparisons
  pairwise_test = "wilcox",
  p_adjust_method = "BH", pairwise_comparisons = list(
    c("fluorinated", "zwitterionic"),
    c("fluorinated", "silicon"),
    c("fluorinated", "PEG")
  )
)
```
