---
title: "curvana: a comprehensive R package for analyzing atomic force microscopy force–distance curves"
subtitle: "Tutorial"
author: Yue (Gary) Zhang, PhD
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    number_sections: false
    highlight: tango
vignette: >
  %\VignetteIndexEntry{curvana tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=TRUE}
knitr::opts_chunk$set(warning = T, message = T,echo = T) 
```


# Introduction

This tutorial demonstrates the full analytical workflow of the **curvana** R package, from raw AFM deflection–displacement curves to transformed force–distance curves and their analytical metrics, such as adhesive force, interaction energy and adhesive and repulsive distance.  

# Preparation 

## Initializing work environment

```{r Initiate, results='hide',warning='False'}
required_packages = c('curvana', 'ggplot2', 'DT', 'dplyr', 'cowplot')
optional_packages = c('ggpubr')

missing_required <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_required) > 0) {
  stop('Please install required packages first: ', paste(missing_required, collapse = ', '))
}

invisible(suppressPackageStartupMessages({
  lapply(required_packages, library, character.only = TRUE)
}))

available_optional <- optional_packages[vapply(optional_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(available_optional) > 0) {
  invisible(suppressPackageStartupMessages({
    lapply(available_optional, library, character.only = TRUE)
  }))
} else {
  message("Optional package 'ggpubr' is not installed; statistical annotation examples will be skipped.")
}

source(system.file("tutorial_helpers.R", package = "curvana"))
devtools::load_all()
```

## Example of raw curve data file

Here we provide an example of an input file. The file contains two columns: "Calc_Ramp_Ex_nm" and "Defl_V_Ex", which are the piezo displacement in nm and cantilever deflection in V for the approach segment of the curve, respectively. The file also has a column named "Calc_Ramp_Rt_nm" and "Defl_V_Rt" for the retract segment. 

Note that the direction of the curve is the contact region at the end (right side) and the non-interaction region at the beginning (left side), which is the default format of Bruker AFM output. However, by convention, the contact region is usually at the beginning of the curve and the non-interaction region is at the end of the curve. To correct this, we can set reverse to True for the corresponding columns (see the next section).

```{r example raw curve}
curve_path <- system.file("extdata", "F_1.spm-F4E1_ForceCurveIndex_0.spm.txt", package = "curvana")
curve_df <- read.table(curve_path, header = TRUE, sep = "\t")
view_df(curve_df,'Example raw curve data')

print(ggplot(curve_df) +
  geom_line(aes(x = Calc_Ramp_Ex_nm, y = Defl_V_Ex, color = "Approach")) +
  geom_line(aes(x = Calc_Ramp_Rt_nm, y = Defl_V_Rt, color = "Retract")) +
  scale_color_manual(values = c("Approach" = "steelblue", "Retract" = "tomato")) +
  labs(x = "Piezo Displacement (nm)", y = "Deflection (V)", color = "Segment") +
  theme_classic())

```

## Load demo data

All examples below use the demo dataset provided in `extdata/` inside the package. To use your own dataset, you can pass the path to the directory holding the data to the "folder" argument. Note that users can increase the threads to enable parallel processing for loading a large folder (e.g., > 5000 curves), which will substantial reduce the computational time.

At this step, we initialize the fdObj named "test" by importing the demo data directory "extdata/" using the "createFdObjFromFolder" function. An fdObj is the core data container in curvana, designed to store atomic force microscopy (AFM) force–distance curve data, including raw curves, transformed data, associated metadata for downstream analysis, and the analytical metrics generated.

By default, createFdObjFromFolder will read all files in the designated folder with suffix ".txt", and expect each file has at least one pair of piezo displacement and cantilever deflection values for approach or retract segments. The name of the columns can be specified by the user, but by default the function will look for columns with standard column names (i.e., Calc_Ramp_Ex_nm, Defl_V_Ex, Calc_Ramp_Rt_nm, Defl_V_Rt) output by Bruker AFM. Users working with non-Bruker input files can specify column headers. Each curve should have the contact region at the beginning and the non-interaction region at the end. If any column needs to be reversed, you can set reverse to True for that column.

If users only have either approach or retract data, they can also use this function to create an fdObj by only specifying the column names for the available segment, and leave the argument related to the missing segment as is. The function will automatically disregard the missing segment and create an fdObj with only the available segment.

```{r load data}
folder = system.file("extdata", package = "curvana")
t = 1
test.input = createFdObjFromFolder(folder = folder, threads = t,
                                  Displacement_Approach = "Calc_Ramp_Ex_nm", # column name for piezo displacement in approach curve
                                  Deflection_Approach = "Defl_V_Ex", # column name for cantilever deflection in approach curve
                                  Displacement_Retract = "Calc_Ramp_Rt_nm", # column name for piezo displacement in retract curve
                                  Deflection_Retract = "Defl_V_Rt", # column name for cantilever deflection in retract curve
                                  reverse_Displacement_Approach = FALSE, # set to FALSE to not reverse the column for piezo displacement in approach curve
                                  reverse_Displacement_Retract = TRUE, # set to TRUE to reverse the column for piezo displacement in retract curve
                                  reverse_Deflection_Approach = TRUE, # set to TRUE to reverse the column for cantilever deflection in approach curve
                                  reverse_Deflection_Retract = FALSE # set to FALSE to not reverse the column for cantilever deflection in retract curve
                                  )

```

Let's take a quick look at the fdObj (named test) we created. We can see that we imported 10 raw curves. Non of them are transformed yet, and "filename" is the only column in the metadata slot.

```{r show init fdobj}
print(test.input)
view_df(test.input@metadata,'metadata')
```

## Manipulate the fdObj object

The demo dataset has measurements made on different surface at different regions. Let's group them in the metadata. 

```{r edit metadata}
test <- test.input
test@metadata <- test@metadata %>% 
  mutate(surface = sub("_.*","",filename),
         surface = case_when(
            surface == 'F' ~ 'fluorinated',
            surface == 'P' ~ 'PEG',
            surface == 'silicon' ~ 'silicon',
            surface == 'Z' ~ 'zwitterionic',
            TRUE ~ surface),
         region = sub(".*?_([0-9]+)\\.spm.*", "\\1", filename)
  )
view_df(test@metadata,'metadata')
test_beforeTransform <- test
```

It is also possible to extract specific samples from an fdObj either by providing their sample names (i.e., row names in the metadata) or by subsetting the object based on conditions applied to metadata columns.

```{r manipulate fdobj with extract}
test.peg = extract(test,by_col = list(surface = "PEG"))
test.silicon = extract(test,by_sample = rownames(test@metadata[test@metadata$surface == 'silicon',]))
```

Just as we might need to subset a fdObj, we might also need to combine two fdObjs (e.g., from different experiments). The "combine" function is designed for this purpose. Note that if there are overlapping samples between the two fdObjs being combined, the "combine" function will return an error.

```{r manipulate fdobj with combine}
test.combine.1 = combineFdObj(test.peg,test.silicon)
#test.combine.2 = combineFdObj(test.paeru,test.paeru2) # should return error
```


# Preview the raw deflection curves

It is important to inspect the raw curves before formal analysis. Here we visualize the raw curves with a heatmap and a scatter plot to check the overall pattern of the curves.

## Heatmap across each sampling event

Here we use the "plot_raw_deflection_heatmap" function to plot a heatmap of min-max scaled deflection values across each sampling event (i.e., each column is a curve, and each row is a data point along the curve). We also annotate columns by "surface" and "region", and set custom colors for each surface type.

This helps us visualize the overall pattern of deflection values across curves, and identify the baseline and sensitivity calibration regions for each curve. 

```{r plot deflection heatmap, fig.height=6,fig.width=10,eval=TRUE}
plot_raw_deflection_heatmap(
  fdobj             = test,
  annotate_columns  = c("surface", "region"),
  annotation_colors = list(
    surface = c(PEG = "#1b9e77", silicon = "#d95f02",
                fluorinated = "#7570b3", zwitterionic = "#e7298a")
  )
)



```


## Scatter plot of deflection curves

We can use the "plot_deflection_curves" function to plot curves in batch. Here, we split curves into sub-panels by column "surface", and color code them by column "region".

```{r plot deflection curves, fig.height=8,fig.width=10}
plot_deflection_curves(test,split_curves_by = 'surface',group_curves_by = 'region',
                       alpha = 0.5,point_size = 0.5,
                       color_map = list(`1` = 'darkred',
                                        `2` = 'orange',
                                        `3` = 'steelblue'))
```


# Curve transformation

One advantage of curvana is that each step of the data analysis is modularized, enabling a flexible and customizable workflow. Curve transformation in curvana involves several steps, including denoise (smoothing), sensitivity calculation, baseline offset calculation, and the application of these parameters to convert raw data into interaction force–separation distance curves. 

## Single curve transformation

### Denoising

To demonstrate how each single curve is transformed, we take a raw curve of the retract section as an example. First, we apply the "denoise_a_curve" function to the raw curve. The function internally applies a Savitzky-Golay filter to smooth the curve, which can help reduce noise and improve the reliability of sensitivity and baseline calculations. The smoothed curve is used for all downstream analysis, while the original raw curve is preserved in the rawCurves slot of the object for reference.

Here we set polynomial order p to 1 and window size n to 3 for the Savitzky-Golay filter, which means the function will fit a linear polynomial to each window of 3 points to smooth the curve, which is equivalent to a simple moving average. Users can adjust these parameters based on the noise level and characteristics of their curves.

```{r single curve denoise}
example1.name = 'silicon_wafer_2.spm-2658_ForceCurveIndex_1.spm'
example1.raw = test_beforeTransform@rawCurves[[example1.name]]
example1.raw = denoise_a_curve(example1.raw, useCurve = "retract", p = 1, n = 3)

ggplot(data = example1.raw) +
  geom_point(aes(x = Calc_Ramp_Rt_nm, y = Defl_V_Rt_original, color = "Original"), size = 1.5) +
  geom_point(aes(x = Calc_Ramp_Rt_nm, y = Defl_V_Rt, color = "Smoothed"), size = 1.5) +
  geom_path(aes(x = Calc_Ramp_Rt_nm, y = Defl_V_Rt, color = "Smoothed"), size = 0.8,alpha=0.5) +
  scale_color_manual(values = c("Original" = "grey70", "Smoothed" = "steelblue")) +
  labs(x = "Calc_Ramp_Rt (nm)", y = "Defl_V_Rt (V)", title = "Denoising a single curve") +
  theme_minimal(base_size = 14) + 
  theme(legend.title=element_blank())

```

To denoise all curves in the object at once, we can run function "denoise_curves." The function internally executes "denoise_a_curve" for each curve separately, and save the smoothed curve in the rawCurves slot of the object (the original, unsmoothed curve will be moved to column "Defl_V_<useCurve>_original"). The original raw curve is preserved in the rawCurves slot of the object for reference. Users can specify the smoothing method and parameters, or choose to skip smoothing if desired.

```{r denoise curves in batch}

test2 = denoise_curves(test_beforeTransform,useCurve = 'approach',p = 1, n = 3,threads = 1)
test2 = denoise_curves(test2,useCurve = 'retract',p = 1, n = 3,threads = 1)

```


### Sensitivity calculation

The function "calc_sensitivity" is designed to automatically to calculate the sensitivity value. Note that the function returns two outputs -- the sensitivity value, and the subset of data points used to calculate the sensitivity value (the second element of the returned list, which is a named dataframe named senscal_segment).

For sensitivity calculation, the function automatically identifies the linear region in the leftmost part of the curve until the data point user set with the argument "end". The function iteratively builds a segment by adding data points from left to right, and performs linear regression on the segment at each step. The iteration continues until the R-squared value of the linear regression drops below a user-defined threshold (R_squared_min) or reaches a user-defined maximum length (end). The slope of the extended segment is then recorded as the sensitivity value for that curve. Users can also set a minimum length for the accepted segment to ensure reliability of sensitivity estimation. 

Here, we set end = 20 to only use the first 20 points of the curve to calculate sensitivity, and set R_squared_min to 0.99 to ensure the selected segment is highly linear. We also set minimum_length to 4, meaning if the final selected segment has fewer than 4 points, the sensitivity calculation will be considered failed and the function will return NA for sensitivity value.

```{r single curve sensitivity analysis}
#example1.name = '180424_fv_pmon2_loc1-18-22.spm'


x = example1.raw$Calc_Ramp_Rt_nm
y = example1.raw$Defl_V_Rt
res.sens = calc_sensitivity(end = 20, x = x, y = y, R_squared_min = 0.99, minimum_length = 4)
example1.raw$`Region used for Sensitivity calculation` = case_when(example1.raw$Calc_Ramp_Rt_nm %in% res.sens$senscal_segment$x ~ "Sensitivity region", TRUE ~ "Other")

ggplot(example1.raw, aes(x = Calc_Ramp_Rt_nm, y = Defl_V_Rt)) +
  geom_point(aes(
    color = `Region used for Sensitivity calculation`  ), size = 1.5) +
  scale_color_manual(
    values = c(
      "Sensitivity region" = "hotpink",
      "Other" = "grey70"
    ),
    name = "Region"
  ) +
  labs(
    x = "Calc_Ramp_Rt (nm)",
    y = "Defl_V_Rt (V)",
    title = "AFM Force Curve Regions"
  ) +
  theme_minimal(base_size = 14)
```

To analyze the sensitivity value for all curves in the object at once, we can run function "analyze_sensitivity." The function internally executes "calc_sensitivity" for each curve separately, and record the sensitivity value of each curve in a metadata column named "sensitivity_V_nm_dir", where dir is the selected curve portion via useCurve argument (either "approach" or "retract"). 

Curves that failed in sensitivity calculation will have an 'NA' in their corresponding cells. The segment of each raw curve used for sensitivity calculation is saved in the object's slot named "senscal_segment".

```{r analyze sensitivity in batch}
test2 = analyze_sensitivity(test_beforeTransform,useCurve = 'approach',threads = 1,end = 20)
test2 = analyze_sensitivity(test2,useCurve = 'retract',threads = 1,end = 20)
view_df(test2@metadata,'metadata with sensitivity values added as columns')
```

### Baseline offsets


Similarly, the offset voltage of each curve can be estimated using the find_baseline() function. We assume that the far-right region of every curve corresponds to the probe being fully detached from the surface and therefore free from interaction forces. The mean of the y values of the last N points (defined by the least_length argument) in this region is calculated as the baseline offset value that will be used in later steps to move the curve up or down to align with the zero-force line (i.e., x axis). Here we set the least length of the curves to 300, as the heatmap above shows that all curves maintain a stable baseline in the region of the last 300 data points. We set slope threshold to 0.01 and standard deviation threshold to 0.01, meaning the curve will be discarded if the last 300 points has a standard deviation of y-value greater than 0.01 or its linear regression has a slope greater than 0.01.

Similar as the "cal_sensitivity" function, "find_baseline" also returns two outputs -- the offset value , and the subset of data points used to calculate the offset value .


```{r single curve baseline calculation}

res.base = find_baseline(x = x, y = y,  sensitivity = res.sens$sensitivity, least_length = 300,slp_threshold = 0.01,std_threshold = 0.01)
example1.raw$`Region used for baseline offset calculation` = case_when(example1.raw$Calc_Ramp_Rt_nm %in% res.base$segment$x ~ "Baseline region", TRUE ~ "Non-baseline region")

ggplot(example1.raw, aes(x = Calc_Ramp_Rt_nm, y = Defl_V_Rt)) +
  geom_point(aes(color = `Region used for baseline offset calculation`), size = 1.5) +
  scale_color_manual(
    values = c(
      "Baseline region" = "orange",
      "Non-baseline region" = "grey70"
    ),
    name = "Region"
  ) +
  labs(
    x = "Calc_Ramp_Ex (nm)",
    y = "Defl_V_Ex (V)",
    title = "AFM Force Curve Regions"
  ) +
  theme_minimal(base_size = 14)
```


To analyze the baseline offset for each curve in an fdObj, we use the analyze_baseline() function. This function internally calls find_baseline() for each curve and records the computed baseline offset values in the metadata columns "baseline_V_dir", along with the length of the baseline segment defined by the user. The portion of each raw curve used for baseline offset estimation are stored in the baseline_segment slot of the object.

Curves can exhibit undulating or unstable baselines. Such abnormal curves—identified by a slope greater than slp_threshold and a standard deviation greater than std_threshold—are considered unreliable and are omitted from subsequent analyses. Curves that fail baseline calculation are assigned NA in these metadata columns. The analyze_baseline function will automatically look up the sensitivity columns generated in analyze_sensitivity. Since baseline estimation requires a valid sensitivity value, any curve that fails sensitivity calculation (i.e., does not have a numeric sensitivity value in the corresponding metadata column) will also fail baseline calculation. 

```{r analyze baseline in batch}
test2 = analyze_baseline(test2,useCurve = 'approach',least_length = 300,slp_threshold = 0.01,std_threshold = 0.01)
test2 = analyze_baseline(test2,useCurve = 'retract', least_length = 300,slp_threshold = 0.01,std_threshold = 0.01)
view_df(test2@metadata,'metadata with baseline values added as columns')
```


## Transformation of a raw curve

Now, with the sensitivity and baseline offset values calculated, we are ready to transform the curve with function transform_a_curve. senscal_seg_x and senscal_seg_y are the x and y values of the segment used for sensitivity calculation, which also will be used for curve transformation. 

```{r single curve transform,fig.height=6,fig.width=16}
example1.transformed = transform_a_curve(x = x, y= y, baseline = res.base$baseline, sensitivity = res.sens$sensitivity,
                             spring_constant = 0.08, senscal_seg_x = res.sens$senscal_segment$x,
                             senscal_seg_y = res.sens$senscal_segment$y)

p1 = ggplot(example1.raw, aes(x = Calc_Ramp_Rt_nm, y = Defl_V_Rt)) +
  geom_point(aes(
    color = case_when(
      `Region used for Sensitivity calculation` == 'Sensitivity region' ~ "Sensitivity region",
      `Region used for baseline offset calculation` == 'Baseline region' ~ "Baseline region",
      TRUE ~ "Other"
    )
  ), size = 1.5) +
  scale_color_manual(
    values = c(
      "Sensitivity region" = "hotpink",
      "Baseline region" = "orange",
      "Other" = "grey70"
    ),
    name = "Region"
  ) +
  labs(
    x = "Calc_Ramp_Ex (nm)",
    y = "Defl_V_Ex (V)",
    title = "Raw curve"
  ) +
  theme_minimal(base_size = 14) + 
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.5)


example1.transformed = cbind(example1.transformed, example1.raw[,c('Region used for Sensitivity calculation','Region used for baseline offset calculation')])
p2 = ggplot(example1.transformed, aes(x = separation_distance_nm, y = force_nN)) +
  geom_point(aes(
    color = case_when(
      `Region used for Sensitivity calculation` == "Sensitivity region" ~ "Sensitivity region",
      `Region used for baseline offset calculation` == "Baseline region" ~ "Baseline region",
      TRUE ~ "Other"
    )
  ), size = 1.5) +
  scale_color_manual(
    values = c(
      "Sensitivity region" = "hotpink",
      "Baseline region" = "orange",
      "Other" = "grey70"
    ),
    name = "Region"
  ) +
  labs(
    x = "Separation distance (nm)",
    y = "force (nN)",
    title = "Transformed curve"
  ) +
  theme_minimal(base_size = 14) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.5)

plot_grid(p1,p2)
```


## Transform curves using the wrapper function "transform_curves"

At this step, we apply the wrapper function transform_curves() to transform all raw curves to calibrated force–distance curves in the fdObj at once. This function streamlines the entire transformation process, by internally calling denoise_curves(), analyze_sensitivity() and analyze_baseline() to denoise raw curves and estimate the sensitivity and baseline values. All arguments accepted by these separate functions can be passed directly to transform_curves(), allowing full control over the analysis parameters. 

__Note: if the sensitivity and baseline columns corresponding to each curve already exist in the metadata, the function will skip the calculation step and directly use the existing values for curve transformation.__

To transform a curve, the spring constant of the AFM probe is required to convert deflection (V) to force (nN). The transform_curves() function accepts either a single numeric value, applied to all curves and stored in metadata$spring_constant, or a metadata column name containing per-curve spring constants. These values are then used with the computed baseline and sensitivity to generate the final force–distance curves.

```{r transform curves approach,warning=TRUE, message=TRUE,results='hold', echo=TRUE}
t = 1
# test_beforeTransform -> test
test = transform_curves(test,
                        denoise_first = TRUE, # whether to apply denoising to raw curves before transformation. Denoising can help improve the reliability of sensitivity and baseline calculations, especially for noisy curves. The smoothed curve will be used for all downstream analysis, while the original raw curve will be preserved in the rawCurves slot of the object for reference.
                        p = 1, n = 3, # parameters for the Savitzky-Golay filter used in curve denoising. p is the polynomial order (default 1, which means linear), and n is the window size (default 3, which means each point will be smoothed by fitting a polynomial to itself and its 2 neighbors). Users can adjust these parameters based on the noise level and characteristics of their curves.
                        
                        spring_constant = 0.08, # spring constant
                        useCurve = 'approach', # analyze the approach curve

                        # baseline determination and filtering
                        least_length = 300, # use the last 300 points as baseline
                        slp_threshold = 0.01, # any curves with the baseline region having slope > 0.01 will be discarded
                        std_threshold = 0.01, # any curves with the baseline region having standard deviation of y values > 0.01 will be discarded

                        # sensitivity calibration
                        intv = 4, # number of data points added per iteration when building the linear sensitivity segment
                        end = 20, # use the first 20 data points at most on the left side of the curve to calibrate sensitivity.
                        R_squared_min = 0.99, # minimum R^2 threshold for accepting a segment as sufficiently linear during sensitivity calculation
                        minimum_length = 4, # minimum number of accumulated points required for a valid sensitivity result; if shorter, sensitivity is reported as NA
                        threads = t)

test = transform_curves(test,
                        denoise_first = TRUE, # whether to apply denoising to raw curves before transformation. Denoising can help improve the reliability of sensitivity and baseline calculations, especially for noisy curves. The smoothed curve will be used for all downstream analysis, while the original raw curve will be preserved in the rawCurves slot of the object for reference.
                        p = 1, n = 3, # parameters for the Savitzky-Golay filter used in curve denoising. p is the polynomial order (default 1, which means linear), and n is the window size (default 3, which means each point will be smoothed by fitting a polynomial to itself and its 2 neighbors). Users can adjust these parameters based on the noise level and characteristics of their curves.
                        
                        spring_constant = 0.08, # spring constant
                        useCurve = 'retract', # analyze the retract curve

                        # baseline determination and filtering
                        least_length = 300, # use the last 300 points as baseline
                        slp_threshold = 0.01, # any curves with the baseline region having slope > 0.01 will be discarded
                        std_threshold = 0.01, # any curves with the baseline region having standard deviation of y values > 0.01 will be discarded

                        # sensitivity calibration
                        intv = 4, # number of data points added per iteration when building the linear sensitivity segment
                        end = 20, # use the first 20 data points at most on the left side of the curve to calibrate sensitivity.
                        R_squared_min = 0.99, # minimum R^2 threshold for accepting a segment as sufficiently linear during sensitivity calculation
                        minimum_length = 4, # minimum number of accumulated points required for a valid sensitivity result; if shorter, sensitivity is reported as NA
                        threads = t)
```

## Soft surface correction

__When measuring on a compliant (soft) surface — such as a hydrogel, cell membrane, or polymer brush — the contact region of the curve reflects a compound response of both cantilever deflection and surface indentation. Using this region to estimate probe sensitivity will therefore underestimate the true probe sensitivity. To correct for this, the probe sensitivity should be calibrated separately on a rigid reference surface (e.g., bare glass or silicon) and then supplied to transform_curves() when processing the soft-surface curves.__

__To do this, set soft = TRUE and provide the externally calibrated sensitivity via probe_sensitivity_external. This can be either a single numeric value (applied to all curves) or the name of a metadata column containing per-curve sensitivity values. When soft = TRUE, transform_curves() uses this external sensitivity to convert deflection (V) to force (nN). Note that the x-axis transformation (piezo displacement to separation distance) is unaffected and always uses the sensitivity estimated from each individual curve, preserving the deformation component in the separation distance.__



```{r use external sensitivity value}

test3 = transform_curves(test,
                        denoise_first = TRUE, # whether to apply denoising to raw curves before transformation. Denoising can help improve the reliability of sensitivity and baseline calculations, especially for noisy curves. The smoothed curve will be used for all downstream analysis, while the original raw curve will be preserved in the rawCurves slot of the object for reference.
                        p = 1, n = 3, # parameters for the Savitzky-Golay filter used in curve denoising. p is the polynomial order (default 1, which means linear), and n is the window size (default 3, which means each point will be smoothed by fitting a polynomial to itself and its 2 neighbors). Users can adjust these parameters based on the noise level and characteristics of their curves.
                        
                        spring_constant = 0.08, # spring constant
                        useCurve = 'retract', # analyze the retract curve

                        # baseline determination and filtering
                        least_length = 300, # use the last 300 points as baseline
                        slp_threshold = 0.01, # any curves with the baseline region having slope > 0.01 will be discarded
                        std_threshold = 0.01, # any curves with the baseline region having standard deviation of y values > 0.01 will be discarded

                        # sensitivity calibration
                        intv = 4, # number of data points added per iteration when building the linear sensitivity segment
                        end = 20, # use the first 20 data points at most on the left side of the curve to calibrate sensitivity.
                        R_squared_min = 0.99, # minimum R^2 threshold for accepting a segment as sufficiently linear during sensitivity calculation
                        minimum_length = 4, # minimum number of accumulated points required for a valid sensitivity result; if shorter, sensitivity is reported as NA
                        threads = t,

                        # soft surface correction
                        soft = TRUE, # whether to apply soft surface correction
                        probe_sensitivity_external = 0.05 # externally calibrated sensitivity value to use for soft surface correction (in V/nm). This can be a single numeric value applied to all curves, or the name of a metadata column containing per-curve sensitivity values.
                        )
```


## Examine the object after transformation

We can see that now we have curves transformed.

```{r examine fdobj after transformation}
print(test)
```

The transformed curves are stored in the retractCurves and approachCurves slots of the fdObj. Note that when the sensitivity or baseline cannot be reliably determined—such as in curves with highly fluctuating probe–surface contact regions or irregular baseline segments—the transformation will fail. In these cases, the corresponding entry in the transformed curve slot (retractCurves or approachCurves) is stored as an empty data frame with zero rows, indicating an unsuccessful transformation.

```{r show retract and approach curve slots}
print(names(test@retractCurves)%>% head()) 
print(names(test@approachCurves)%>% head())
```

The calculated sensitivity, baseline, and input spring constant is now saved as columns in the metadata. Note the curves with baseline/sensitivity undetermined will be missing in the corresponding cell.

```{r show metadata after transformation}
view_df(test@metadata,'metadata after curve transformation')
```

## Plot transformed curves

To visualize the transformed curves, we can use the "plot_fd_curves" function. Here we plot the approach and retract curves in separate panels, and color code them by column "surface".

```{r plot transformed curves, fig.height=12,fig.width=10}
plot_fd_curves(test,group_curves_by = 'region', split_curves_by = 'surface',
               point_alpha = 0.5,line_alpha = 0.3, point_size = 0.5,
               color_map = list(`1` = 'darkred',
                                `2` = 'orange',
                                `3` = 'steelblue'))
```

We can see that Fluorinated surface has the largest adhesion, which occurs mostly immediately after pulling probe off the surface. The other surfaces also have adhesion, but the magnitude of adhesion is smaller. All curves passed the transformation step have a baseline that is well aligned with the zero-force line with a relatively small fluctuation, which is important for accurate calculation of analytical metrics in later steps.


# Analytical metrics calculation

After curve transformation, we can calculate analytical metrics such as adhesive force, interaction energy and distance.

## Baseline noise-level determination

### Wrapper function for batch analysis

Before calculating analytical metrics, it is important to determine the baseline noise level so we can set a threshold to exclude insignificant fluctuations in metric calculations. The function "analyze_a_curve_noise" estimates a lower and upper noise band from baseline force values. For batch analysis, "analyze_curves_noise" applies this to all transformed curves in an fdObj and writes the results into metadata columns `noiseBand_low_nN` and `noiseBand_high_nN` (one set for approach and one set for retract).

Supported noise-band determination methods are standard deviation (`sd`), median absolute deviation (`mad`), quantile (`quantile`), and fixed (`fixed`). The `multiplier` parameter is used to multiply the chosen noise metric to define the noise band. The noise-band are then reused by interaction-distance and interaction energy analysis, so thresholds are determined once and applied consistently across curves.

Here we set noise band as the region ranging from the minimum and maximum of the force in the baseline region.

__Note: set noise band multiplier to 0 to not apply noise band thresholding.__
```{r calculate baseline noise level}
test2 = analyze_curves_noise(test,
                            useCurve = 'approach', # calculate noise band for approach curves,
                            baseline_span = "automatic", # number of data points at the end of the curve to use as baseline region for noise level calculation. Here we set it to "automatic", which means the function will use the baseline_span column in the metadata.
                            threshold_method = 'quantile', # method to determine noise band. Here we use quantile, which means the noise band will be defined by the quantiles of the baseline force values. The specific quantiles are determined by quantile_low and quantile_high parameters.
                            quantile_low = 0, # lower quantile for noise band. Here we set it to 0, meaning the lower bound of the noise band will be the minimum force value in the baseline region.
                            quantile_high = 1, # upper quantile for noise band. Here we set it to 1, meaning the upper bound of the noise band will be the maximum force value in the baseline region.
                            multiplier = 1 # multiplier for the noise metric. We set multiplier to 1 to directly use the min-max range of the force in baseline region without further scaling.
                            ) 
test2 = analyze_curves_noise(test2, useCurve = 'retract',
                            threshold_method = 'quantile',quantile_low = 0,quantile_high = 1,
                            multiplier = 1)
view_df(test2@metadata,'metadata with baseline noise level values added as columns')
```

### Single curve analysis

To demonstrate how the baseline noise level is calculated for a single curve, we use the above transformed retract curve as an example. The function "analyze_a_curve_noise" estimates the lower and upper noise bands based on the specified method and multiplier, using the baseline region of the curve, as discussed above. 

Here we use the same parameters as in the batch analysis, which means the noise band will be the region ranging from the minimum and maximum of the force in the baseline region. The function returns a named vector containing the calculated lower and upper noise band values.

```{r single curve noise level calculation,fig.height=6,fig.width=8}
res.noise = analyze_a_curve_noise(example1.transformed, baseline_span = 300, threshold_method = 'quantile', multiplier = 1,quantile_low = 0,quantile_high = 1)
print(res.noise)

```


## Adhesive force

### Wrapper function for batch analysis

The adhesive force is calculated as the highest negative force value in a curve, which represents the maximum adhesive force between the AFM probe and the sample surface. The "analyze_curves_adhesive_force" function identifies this value and the separation distance at which it occurs, and records the force in the metadata column named "adhesive_force_nN" and the distance in column "adhesive_sep_nm". Curves that fail transformation will have NA in this column.

__Note: If the minimum force of the curve is positive, the adhesive force will be reported as 0, and the associated separation distance will be NA. If the minimum force is negative, the adhesive force will be reported as the absolute value of the force.__

```{r calculate adhesive force}
test2 = analyze_curves_adhesive_force(test2,useCurve = 'approach',threads = 1)
test2 = analyze_curves_adhesive_force(test2,useCurve = 'retract',threads = 1)
view_df(test2@metadata,'metadata with adhesive force values added as columns')   
```

### Single curve analysis

The function "analyze_a_curve_adhesive_force" identifies the loweset negative force value in the curve and the corresponding separation distance, and returns these values in a named vector.

```{r single curve adhesive force calculation}
res.af = analyze_a_curve_adhesive_force(example1.transformed)
print(res.af)
```

## Interaction energy

In AFM force spectroscopy, interaction energy is quantified by calculating the area under the curve. The analyze_a_curve_area() function computes adhesive energy (area below the negative noise threshold and above the curve inside Quadrant IV, representing bond-breaking energy) and repulsive energy (area above the positive noise threshold and under the curve inside Quadrant I). A noise band is applied to filter out thermal fluctuations and instrumental noise, which would otherwise artificially inflate energy measurements. By excluding force values within the noise band, the function ensures that only biologically or physically meaningful interactions are counted, yielding more accurate energies.

### Wrapper function for batch analysis

The wrapper function "analyze_curves_energy" applies this analysis to all curves in an fdObj, and records the calculated adhesive and repulsive energy in metadata (columns adhesive_energy_fJ and repulsive_energy_fJ, one set for retract and one set for approach curves). Curves that fail transformation will have NA in these columns. The function will automatically look up the metadata in the object to find the pre-determined noise band.

Note that for curves that are over-corrected, which have a portion of curve near the contact region extends to the negative side of y axis (i.e., having negative separation distance), the function will automatically correct all negative x to zero before energy calculation

```{r calculate interaction energy}
test2 = analyze_curves_energy(test2,useCurve = 'approach',threads = 1)
test2 = analyze_curves_energy(test2,useCurve = 'retract',threads = 1)

view_df(test@metadata,'metadata with energy values added as columns') 
```

### Single curve analysis

The function "analyze_a_curve_energy" computes the adhesive energy and repulsive energy based on the area under the curve, applying a noise cutoff to exclude insignificant fluctuations. To avoid counting noise as part of the interaction energy, we set noise cutoff with the computed noise band. 

```{r single curve interaction energy calculation, fig.height=6, fig.width=8}
analyze_a_curve_area(example1.transformed, noiseBand_low = res.noise[['noiseBand_low']], noiseBand_high = res.noise[['noiseBand_high']])
```

## Interaction distance

The interaction distance is calculated as the separation distance at which the force curve first crosses the noise threshold, indicating the start or end of significant interaction between the AFM probe and the sample surface. Specifically, by configuring the function to scan from different directions (left-to-right or right-to-left) and to look for either positive or negative excursions, users can identify different types of interaction distances. 


### Rupture/adhesion distance

For example, for rupture events (negative-force excursions), the function can be set to scan from right to left and identify the first point where the curve dips below the negative noise threshold, which indicates the start of significant adhesion.

Specifically, the function will scan the curve from right to left (x_direction = "left"), with excluding the baseline_region (defined by baseline_span), to look for the first negative-force excursions (y_direction = "negative") with at least 3 consecutive points below the threshold (min_consecutive = 3).

```{r single curve rupture distance}
res.rupturedis = analyze_a_curve_interaction_distance(example1.transformed,
                                                      y_direction = "negative", 
                                                      x_direction = "left",
                                                      baseline_span = 200,
                                                      noiseBand_low = res.noise[['noiseBand_low']],
                                                      noiseBand_high = res.noise[['noiseBand_high']],
                                                      min_consecutive = 3
                                                      )

print(res.rupturedis)

```

To analyze rupture distance for all curves in an fdObj, we can use the wrapper function "analyze_curves_interaction_distance" with the same parameters. The calculated rupture distance will be recorded in a metadata column named "rupture_distance_nm". Curves that fail transformation will have NA in this column.

```{r rupture distance for fdobj}
test2 = analyze_curves_interaction_distance(test2,
                                           useCurve = 'retract', # analyze retract curve
                                           y_direction = "negative", x_direction = "left",
                                           baseline_span = 'automatic', # use the baseline span values in metadata column "baseline_span_retract" for each curve
                                            min_consecutive = 3
                                            )

view_df(test2@metadata,'metadata with rupture length')
```

### Repulsive distance

For repulsive interactions (positive-force excursions), the function can be set to scan from left to right and identify the last point before the curve cross the positive noise threshold to enter the noise band region from the repulsive force region, which indicates the end of significant repulsion.

Specifically, the function scans the curve from left to right (x_direction = "right") up to the beginning of the baseline region (as defined by baseline_span). It identifies the last data point where the curve first crosses the positive noise threshold—entering the noise band from the repulsive force region (y_direction = "positive"). Because the curve is expected to begin in the repulsive region, min_consecutive is set to 1, meaning that a single point above the positive noise threshold is sufficient to mark the start of the repulsive region.


```{r single curve repulsive distance}
res.repuldis = analyze_a_curve_interaction_distance(example1.transformed,
                                                      y_direction = "positive", 
                                                      x_direction = "right",
                                                      baseline_span = 200,
                                                      noiseBand_low = res.noise[['noiseBand_low']],
                                                      noiseBand_high = res.noise[['noiseBand_high']],
                                                      min_consecutive = 1
                                                      )

print(res.repuldis)

```

To analyze repulsive distance for all curves in an fdObj, we can use the wrapper function "analyze_curves_interaction_distance" with the same parameters. The calculated repulsive distance will be recorded in a metadata column named "repulsive_distance_nm". Curves that fail transformation will have NA in this column.

```{r repulsive distance for fdobj}
test2 = analyze_curves_interaction_distance(test2,
                                           useCurve = 'approach', # analyze approach curve
                                           y_direction = "positive", x_direction = "right",
                                           baseline_span = 'automatic', # use the baseline span values in metadata column "baseline_span_approach" for each curve
                                           min_consecutive = 1
                                          )

view_df(test2@metadata,'metadata with repulsive distance')
```

## All-in-one analytical wrapper

You can run all analytical-metric steps described above in one call using `analyze_curves_all_analytical_metrics()`. This function combines all the above analyses into a single streamlined workflow, allowing you to compute noise levels, adhesive forces, interaction energies, and interaction distances for all curves in an fdObj with one command. By setting the appropriate toggles, you can customize which metrics to calculate and how to apply noise thresholds, providing a comprehensive analysis of your AFM force spectroscopy data in a single step.

```{r run all analytical metrics with toggles}
test = analyze_curves_all_analytical_metrics(
  test, # the object
  useCurve = "both", # run analysis for both approach and retract 
  threads = 1, 

  # noise band parameters
  noise_baseline_span = "automatic", # use the baseline section previously defined in curve transformation for noise band calculation
  noise_threshold_method = "quantile", # use quantile method to define the noise band
  noise_quantile_low = 0.00, # define the lower end of noise band as minimum y value of the baseline section
  noise_quantile_high = 1, # define the upper end of noise band as maximum y value of the baseline section
  noise_multiplier = 1, # multiply the noise band by 1 to directly use the min-max range of the force in baseline region without further scaling.

  # adhesive force
  analyze_adhesive_force = TRUE, # whether to analyze adhesive force

  # adhesive energy and repulsive energy
  analyze_energy = TRUE, # whether to analyze interaction energy (repulsive and adhesive)

  # rupture (adhesive) distance
  analyze_rupture_distance = TRUE, # whether to analyze adhesive/rupture distance
  analyze_rupture_distance_baseline_span = "automatic", # use the baseline section previously defined in curve transformation for rupture distance analysis
  analyze_rupture_distance_x_direction = "left", # scan from right to left to find the first data point at which a curve enters the adhesive region from the noise band region
  analyze_rupture_distance_min_consecutive = 3, # to reduce false positives from random fluctuations, require at least 3 consecutive points below the lower noise-band threshold for classifying the curve as entering the adhesive region.

  # repulsive distance
  analyze_repulsive_distance = TRUE, # whether to analyze repulsive distance
  analyze_repulsive_distance_baseline_span = "automatic", # use the baseline section previously defined in curve transformation for repulsive distance analysis
  analyze_repulsive_distance_x_direction = "right", # scan from left to right to find the last data point before a curve enters the noise band region from the repulsive region
  analyze_repulsive_distance_min_consecutive = 1 # because the curve is expected to start from the repulsive region, we set min_consecutive to 1, meaning that a single point above the positive noise threshold is sufficient to mark the start of the repulsive region.
)
```


# Visualization of analytical metrics

plot_a_curve_metrics() can be used to visualize the calculated analytical metrics for a single curve.

```{r single curve plot metrics,fig.height=12, fig.width=22}
plot_a_curve_metrics(
  fdobj = test,
  curve_name = example1.name,
  useCurve = "retract", # which curve to plot, can be "approach" or "retract".
  plot_raw = T, # whether to plot the raw curve together with the transformed curve. 
  base_size = 15,
  annotation_text_size = 5
)

```

# Save all raw and transformed curves with analytical metrics annotated (optional)

```{r plot all metrics for a curve,fig.height=8, fig.width=16,eval=F}
plot_all_curve_metrics(
  fdobj = test, # the fdObj to plot
  useCurve = "retract", # plot the retract curve
  plot_raw = T, # whether to plot the raw curve
  destination_folder = '../test_plot_metrics_denoised', # the path to save the plots
  format = 'png', width = 18, height = 9, # plot format and dimensions,
  threads = 8 # number of threads to use for parallel processing (set to 8 to use 8 cores, which will significantly speed up the plotting process for a large dataset
)
plot_all_curve_metrics(
  fdobj = test, # the fdObj to plot
  useCurve = "approach", # plot the approach curve
  plot_raw = T, # whether to plot the raw curve
  destination_folder = '../test_plot_metrics_denoised', # the path to save the plots
  format = 'png', width = 18, height = 9, # plot format and dimensions
  threads = 8 # number of threads to use for parallel processing (set to 8 to use 8 cores, which will significantly speed up the plotting process for a large dataset
)
```

# Summary statistics of analytical metrics

## PCA biplot of selected metrics

Dimensional reduction is first used to visualize the overall pattern of how curves cluster based on their analytical metrics. The function "plot_pca_biplot" performs principal component analysis (PCA) on the selected metrics and plots the scores of samples in the PCA space, colored by a specified metadata variable (e.g., surface type). The function also plots the loadings of the selected features as arrows, indicating how each metric contributes to the principal components and the clustering pattern of samples.

```{r pca biplot of metrics, fig.height=6, fig.width=7}
pca_features <- c(
  "adhesive_force_nN_retract",
  "adhesive_energy_aJ_retract",
  "rupture_distance_nm_retract",
  "repulsive_energy_aJ_approach",
  "repulsive_distance_nm_retract"
)

plot_pca_biplot(
  df = test@metadata,
  include_columns = pca_features,
  color_by = "surface",
  color_map = c(
    PEG = "#1b9e77",
    silicon = "#d95f02",
    fluorinated = "#7570b3",
    zwitterionic = "#e7298a"
  ),
  arrow_scale = 1.1,
  show_feature_labels = TRUE
)
```

From the PCA biplot, we can see the fluorinated surface is well separated from the other three surfaces, which is consistent with our observation of the curves and the fact that fluorinated surface has the largest adhesion. The biplot also shows that adhesive force, adhesive energy and rupture distance are the main contributors to PC1, which separates fluorinated surface from the others, while repulsive energy and repulsive distance contribute more to PC2, which largely separates zwitterionic surface from PEG and silicon surfaces. Adhesive force, adhesive energy and rupture distance are positively correlated with each other, which is consistent with the fact that they all reflect the strength of adhesion. Repulsive energy and repulsive distance are also positively correlated, which makes sense because a stronger repulsion would typically extend over a longer distance.


## Scaled feature heatmap with sample annotations

The function `plot_complex_heatmap()` creates a heatmap of selected analytical metrics across samples, with options for clustering and annotation. In this example, we select the same metrics used for PCA plot to visualize their relative values across samples. The heatmap is scaled by z-score to show how each sample deviates from the mean for each metric, allowing for easy comparison across different metrics with different units and ranges. The samples are annotated by surface type and region, with a specified color map for the surface types. Clustering is applied to rows (samples) to reveal patterns of similarity based on the selected metrics.

```{r complex heatmap of metrics, fig.height=5, fig.width=8}
  plot_complex_heatmap(
    df = test@metadata,
    include_columns = pca_features,
    annotate_columns = c("surface", "region"),
    annotation_colors = list(
      surface = c(
        PEG = "#1b9e77",
        silicon = "#d95f02",
        fluorinated = "#7570b3",
        zwitterionic = "#e7298a"
      )
    ),
    cluster_rows = TRUE,
    cluster_columns = FALSE,
    show_row_dend = TRUE,
    show_column_dend = FALSE,
    heatmap_name = "z-score",
    show_column_names = FALSE
  )
```

## Violin plot of selected metrics by surface

`plot_metric_violin()` can be used to visualize the distribution of selected metrics across different surfaces. Here we plot the violin plot of adhesive force, adhesive energy and rupture distance of the retract curves, as well as the repulsive energy and repulsive distance of approach curves. The option "show_whisker_box" is set to TRUE to show the median and interquartile range of the distribution.

```{r summary statistics of metrics, fig.height=6, fig.width=12}
f1 = plot_metric_violin(test@metadata, metric_name = "adhesive_force_nN_retract", group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"),
show_whisker_box = T)
f2 = plot_metric_violin(test@metadata, metric_name = "adhesive_energy_aJ_retract",log10=T,  group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"),
show_whisker_box = T)
f3 = plot_metric_violin(test@metadata, metric_name = "repulsive_energy_aJ_approach", group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"),
show_whisker_box = T)
f4 = plot_metric_violin(test@metadata, metric_name = "rupture_distance_nm_retract", log10=T, group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"),
show_whisker_box = T)
f5 = plot_metric_violin(test@metadata, metric_name = "repulsive_distance_nm_approach", log10=T, group_by = "surface", color_map = c(PEG = "#1b9e77", silicon = "#d95f02", fluorinated = "#7570b3", zwitterionic = "#e7298a"),
show_whisker_box = T)

plot_grid(f1,f2,f3,f4,f5, nrow = 2)


```

## Violin plot with statistical annotations

`plot_metric_violin()` also has options to perform statistical tests and annotate the significance on the plot. Here we perform Kruskal-Wallis test to compare the means of adhesive force across different surfaces, and perform pairwise Wilcoxon tests for post-hoc analysis. The p-values are adjusted using Benjamini-Hochberg method to control for false discovery rate.

```{r summary statistics with tests, eval=requireNamespace("ggpubr", quietly = TRUE)}
plot_metric_violin(
  df = test@metadata,
  metric_name = "adhesive_force_nN_retract",
  group_by = "surface",
  color_by = "surface",
  color_map = c(
    PEG = "#1b9e77",
    silicon = "#d95f02",
    fluorinated = "#7570b3",
    zwitterionic = "#e7298a"
  ),
  global_test = "kruskal",
  pairwise_test = "wilcox",
  p_adjust_method = "BH"

)
```

The function also allows users to specify which pairwise comparisons to perform, so that only relevant comparisons are made and the plot is not overcrowded with too many significance annotations. Here we specify to compare fluorinated surface with each of the other three surfaces.

```{r summary statistics with tests2, eval=requireNamespace("ggpubr", quietly = TRUE)}
plot_metric_violin(
  df = test@metadata,
  metric_name = "adhesive_force_nN_retract",
  group_by = "surface",
  color_by = "surface",
  color_map = c(
    PEG = "#1b9e77",
    silicon = "#d95f02",
    fluorinated = "#7570b3",
    zwitterionic = "#e7298a"
  ),
  global_test = "none", # do not perform global test, only do pairwise comparisons
  pairwise_test = "wilcox",
  p_adjust_method = "BH", pairwise_comparisons = list(
    c("fluorinated", "zwitterionic"),
    c("fluorinated", "silicon"),
    c("fluorinated", "PEG")
  )
)
```