---
title: "curvana: a comprehensive R package for analyzing atomic force microscopy force–distance curves"
subtitle: "Tutorial"
author: Yue (Gary) Zhang, PhD
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    number_sections: false
    highlight: tango
vignette: >
  %\VignetteIndexEntry{curvana vignette: full workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=TRUE}
knitr::opts_chunk$set(warning = T, message = T,echo = T) 
```


# Introduction

This tutorial demonstrates the full analytical workflow of the **curvana** R package, from raw AFM deflection-displacement curves to transformed force-distance curves and their analytical metrics, such as adhesive force, interaction energy, and adhesive and repulsive distances.  

# Preparation 

## Initializing the work environment

```{r Initiate, results='hide',warning='False'}
#required_packages = c('curvana', 'ggplot2', 'DT', 'dplyr', 'cowplot')
required_packages = c('curvana', 'ggplot2', 'dplyr', 'cowplot')
optional_packages = c('DT', 'ggpubr')

missing_required <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_required) > 0) {
  stop('Please install required packages first: ', paste(missing_required, collapse = ', '))
}

invisible(suppressPackageStartupMessages({
  lapply(required_packages, library, character.only = TRUE)
}))

available_optional <- optional_packages[vapply(optional_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(available_optional) > 0) {
  invisible(suppressPackageStartupMessages({
    lapply(available_optional, library, character.only = TRUE)
  }))
} else {
  message("Optional package 'ggpubr' is not installed; statistical annotation examples will be skipped.")
}

source(system.file("tutorial_helpers.R", package = "curvana"))

```

## Example of raw curve data file

Below is an example input file. It contains two columns, "Calc_Ramp_Ex_nm" and "Defl_V_Ex", which represent the piezo displacement in nm and cantilever deflection in V for the approach segment of the curve, respectively. The file also includes the columns "Calc_Ramp_Rt_nm" and "Defl_V_Rt" for the retract segment. 

Note that in the default Bruker AFM output format, the contact region is at the end of the curve (right side) and the non-interaction region is at the beginning (left side). By convention, however, the contact region is usually placed at the beginning of the curve and the non-interaction region at the end. To correct this, set `reverse = TRUE` for the corresponding columns, as shown in the next section.

```{r example raw curve}
curve_path <- system.file("extdata", "F_1.spm-F4E1_ForceCurveIndex_0.spm.txt", package = "curvana")
curve_df <- read.table(curve_path, header = TRUE, sep = "\t")
view_df(curve_df,'Example raw curve data')

print(ggplot(curve_df) +
  geom_line(aes(x = Calc_Ramp_Ex_nm, y = Defl_V_Ex, color = "Approach")) +
  geom_line(aes(x = Calc_Ramp_Rt_nm, y = Defl_V_Rt, color = "Retract")) +
  scale_color_manual(values = c("Approach" = "steelblue", "Retract" = "tomato")) +
  labs(x = "Piezo Displacement (nm)", y = "Deflection (V)", color = "Segment") +
  theme_classic())

```

## Load demo data

All examples below use the demo dataset provided in `extdata/` within the package. This folder contains approximately 100 raw curve files exported from Bruker NanoScope Analysis with both approach and retract segments included. To use your own dataset, pass the path to the directory containing the data to the `folder` argument. Users can increase `threads` to enable parallel processing when loading a large folder, for example, more than 5000 curves, which can substantially reduce computation time.

In this step, we initialize an fdObj by importing the demo data directory `extdata/` with `createFdObjFromFolder()`. An fdObj is the core data container in curvana. It stores atomic force microscopy (AFM) force-distance curve data, including raw curves, transformed data, associated metadata for downstream analysis, and the analytical metrics generated.

### Bruker/Veeco-style ASCII

Use `createFdObjFromFolder()` for generic Bruker/Veeco-style column-separated files. By default, the function reads all `.txt` files (or `.csv` if `suffix = ".csv"`) and removes lines starting with `#` or `"` before import. It expects at least one valid segment (approach or retract) with displacement and deflection columns. The default Bruker column names are `Calc_Ramp_Ex_nm`, `Defl_V_Ex`, `Calc_Ramp_Rt_nm`, and `Defl_V_Rt`, but these can be replaced with custom headers for other exports. If contact and non-interaction regions are in the opposite orientation, use the `reverse_*` arguments to correct direction. If users have only approach or retract data, they can still use this function to create an fdObj by specifying column names only for the available segment and leaving the arguments for the missing segment unchanged. The function will automatically ignore the missing segment and create an fdObj containing only the available data.



```{r load data}
folder = system.file("extdata", package = "curvana")
t = 1
test.input = createFdObjFromFolder(folder = folder, threads = t,
                                  Displacement_Approach = "Calc_Ramp_Ex_nm", # column name for piezo displacement in the approach curve
                                  Deflection_Approach = "Defl_V_Ex", # column name for cantilever deflection in the approach curve
                                  Displacement_Retract = "Calc_Ramp_Rt_nm", # column name for piezo displacement in the retract curve
                                  Deflection_Retract = "Defl_V_Rt", # column name for cantilever deflection in the retract curve
                                  reverse_Displacement_Approach = FALSE, # set to FALSE to keep the piezo displacement column for the approach curve as is
                                  reverse_Displacement_Retract = TRUE, # set to TRUE to reverse the piezo displacement column for the retract curve
                                  reverse_Deflection_Approach = TRUE, # set to TRUE to reverse the cantilever deflection column for the approach curve
                                  reverse_Deflection_Retract = FALSE # set to FALSE to keep the cantilever deflection column for the retract curve as is
                                  )

```

### JPK-style ASCII

For JPK ASCII exports, use the dedicated importer `createFdObjFromJPKFolder()` instead of `createFdObjFromFolder()`. The JPK importer automatically detects approach/retract segments, parses spring constant and sensitivity, and converts force data back to Voltage when needed. Users are encouraged to directly import raw deflection voltage in analysis.

Specifically, `createFdObjFromJPKFolder()`:

- Detects approach and retract segments automatically
- Parses sensitivity and spring constant for each segment; if deflection is in force units (N), converts it back to voltage
- Converts distance to nm, applies optional segment reversal, and records imported parameters plus datapoint counts in metadata

```{r load data jpk example, eval=FALSE}
# For JPK ASCII data, use the JPK-specific importer:
jpk_folder <- system.file("extdata_jpk", package = "curvana")

test.input.jpk <- createFdObjFromJPKFolder(
  folder = jpk_folder,
  threads = 1,
  height_col = "height",
  deflection_col = "vDeflection"
)
```

Let's take a quick look at the fdObj we created. We can see that 101 raw curves were imported. None of them have been transformed yet, and `filename` is the only column in the metadata slot.

```{r show init fdobj}
print(test.input)
view_df(test.input@metadata,'metadata')
```



## Manipulate the fdObj object

The demo dataset contains measurements made on different surfaces and at different regions. Let's group them in the metadata. 

```{r edit metadata}
test <- test.input
test@metadata <- test@metadata %>% 
  mutate(surface = sub("_.*","",filename),
         surface = case_when(
            surface == 'F' ~ 'Group_A',
            surface == 'P' ~ 'Group_B',
            surface == 'silicon' ~ 'Group_C',
            surface == 'Z' ~ 'Group_D',
            TRUE ~ surface),
         region = sub(".*?_([0-9]+)\\.spm.*", "\\1", filename)
  )
view_df(test@metadata,'metadata')
test_beforeTransform <- test
```

It is also possible to extract specific samples from an fdObj either by providing their sample names, that is, row names in the metadata, or by subsetting the object based on conditions applied to metadata columns.

```{r manipulate fdobj with extract}
test.group_b = extract(test,by_col = list(surface = "Group_B"))
test.group_c = extract(test,by_sample = rownames(test@metadata[test@metadata$surface == 'Group_C',]))
```

Just as we might need to subset an fdObj, we might also need to combine two fdObjs, for example, from different experiments. The `combineFdObj()` function is designed for this purpose. If there are overlapping samples between the two fdObjs being combined, `combineFdObj()` will return an error.

```{r manipulate fdobj with combine}
test.combine.1 = combineFdObj(test.group_b,test.group_c)
#test.combine.2 = combineFdObj(test.paeru,test.paeru2) # should return error
```


# Preview the raw deflection curves

It is important to inspect the raw curves before formal analysis. Here, we visualize them with a heatmap and a scatter plot to examine their overall patterns.

## Deflection vs data point index

Here we use `plot_deflection_curves_by_index()` to plot raw deflection values against the row index (data point number) for each curve. This visualization helps us:
- See how many data points are in each curve
- Verify that the data is oriented in the correct direction (contact region at the beginning, non-interaction region at the end)
- Help us with selecting the appropriate regions for baseline calibration and sensitivity calculation in the next step.

We split curves into subpanels by `surface` and color-code them by `region`.

```{r plot deflection by index, fig.height=6,fig.width=10,eval=TRUE}
plot_deflection_curves_by_index(test,split_curves_by = 'surface',group_curves_by = 'region',
                       alpha = 0.5,point_size = 0.5,
                       color_map = list(`1` = 'darkred',
                                        `2` = 'orange',
                                        `3` = 'steelblue'))



```



## Scatter plot of deflection curves

We can use `plot_deflection_curves()` to plot curves in batch. Here, we split curves into subpanels by `surface` and color-code them by `region`.

```{r plot deflection curves, fig.height=8,fig.width=10}
plot_deflection_curves(test,split_curves_by = 'surface',group_curves_by = 'region',
                       alpha = 0.5,point_size = 0.5,
                       color_map = list(`1` = 'darkred',
                                        `2` = 'orange',
                                        `3` = 'steelblue'))
```


# Curve transformation

One advantage of curvana is that each step of the data analysis is modularized, enabling a flexible and customizable workflow. Curve transformation in curvana involves several steps, including denoising, sensitivity calculation, baseline offset calculation, and application of these parameters to convert raw data into interaction force-separation distance curves. 

## Single curve transformation

### Denoising

To demonstrate how an individual curve is transformed, we use a raw retract curve as an example. First, we apply `denoise_a_curve()` to the raw curve. The function internally applies a Savitzky-Golay filter to smooth the curve, which can help reduce noise and improve the reliability of sensitivity and baseline calculations. The smoothed curve is used for downstream analysis, while the original raw curve is preserved in the `rawCurves` slot of the object for reference.

Here we set the polynomial order `p` to 1 and the window size `n` to 3 for the Savitzky-Golay filter. This means the function fits a linear polynomial to each window of 3 points to smooth the curve, which is equivalent to a simple moving average. Users can adjust these parameters based on the noise level and characteristics of their curves.

```{r single curve denoise}
example1.name = 'silicon_wafer_2.spm-2658_ForceCurveIndex_1.spm'
example1.raw = test_beforeTransform@rawCurves[[example1.name]]
example1.raw = denoise_a_curve(example1.raw, useCurve = "retract", p = 1, n = 3)

ggplot(data = example1.raw) +
  geom_point(aes(x = Calc_Ramp_Rt_nm, y = Defl_V_Rt_original, color = "Original"), size = 1.5) +
  geom_point(aes(x = Calc_Ramp_Rt_nm, y = Defl_V_Rt, color = "Smoothed"), size = 1.5) +
  geom_path(aes(x = Calc_Ramp_Rt_nm, y = Defl_V_Rt, color = "Smoothed"), size = 0.8,alpha=0.5) +
  scale_color_manual(values = c("Original" = "grey70", "Smoothed" = "steelblue")) +
  labs(x = "Calc_Ramp_Rt (nm)", y = "Defl_V_Rt (V)", title = "Denoising a single curve") +
  theme_minimal(base_size = 14) + 
  theme(legend.title=element_blank())

```

To denoise all curves in the object at once, we can run `denoise_curves()`. The function internally executes `denoise_a_curve()` for each curve separately and saves the smoothed curve in the `rawCurves` slot of the object. The original, unsmoothed curve is moved to the `Defl_V_<useCurve>_original` column. Users can specify the smoothing method and parameters, or choose to skip smoothing if desired.

```{r denoise curves in batch}
test2 = denoise_curves(test_beforeTransform,useCurve = 'approach',p = 1, n = 3,threads = 1)
test2 = denoise_curves(test2,useCurve = 'retract',p = 1, n = 3,threads = 1)
```


### Sensitivity calculation

The function `calc_sensitivity()` is designed to calculate the sensitivity value automatically. It returns two outputs: the sensitivity value and the subset of data points used to calculate it. The second element of the returned list is a named data frame called `senscal_segment`.

For sensitivity calculation, the function automatically identifies the linear region in the leftmost part of the curve up to the data point specified by the `end` argument. It iteratively builds a segment by adding data points from left to right and performs linear regression on the segment at each step. The iteration continues until the $R^2$ value of the linear regression drops below the user-defined threshold `R_squared_min` or reaches the user-defined maximum length `end`. The slope of the final segment is then recorded as the sensitivity value for that curve. Users can also set a minimum accepted segment length `minimum_length` to ensure reliable sensitivity estimation. 

Here, we set `end = 20` to use only the first 20 points of the curve for sensitivity calculation, and `R_squared_min = 0.99` to ensure that the selected segment is highly linear. We also set `minimum_length = 4`, meaning that if the final selected segment has fewer than 4 points, the sensitivity calculation is considered to have failed and the function returns `NA`.

```{r single curve sensitivity analysis}
x = example1.raw$Calc_Ramp_Rt_nm
y = example1.raw$Defl_V_Rt
res.sens = calc_sensitivity(end = 20, x = x, y = y, R_squared_min = 0.99, minimum_length = 4)
example1.raw$`Region used for Sensitivity calculation` = case_when(example1.raw$Calc_Ramp_Rt_nm %in% res.sens$senscal_segment$x ~ "Sensitivity region", TRUE ~ "Other")

ggplot(example1.raw, aes(x = Calc_Ramp_Rt_nm, y = Defl_V_Rt)) +
  geom_point(aes(
    color = `Region used for Sensitivity calculation`  ), size = 1.5) +
  scale_color_manual(
    values = c(
      "Sensitivity region" = "hotpink",
      "Other" = "grey70"
    ),
    name = "Region"
  ) +
  labs(
    x = "Calc_Ramp_Rt (nm)",
    y = "Defl_V_Rt (V)",
    title = "AFM Force Curve Regions"
  ) +
  theme_minimal(base_size = 14)
```

To analyze sensitivity values for all curves in the object at once, we can run `analyze_sensitivity()`. The function internally executes `calc_sensitivity()` for each curve separately and records the sensitivity value of each curve in a metadata column named `sensitivity_V_nm_dir`, where `dir` is the selected curve portion specified by the `useCurve` argument, either `approach` or `retract`. 

Curves that fail sensitivity calculation will have `NA` in their corresponding cells. The segment of each raw curve used for sensitivity calculation is saved in the object's `senscal_segment` slot.

```{r analyze sensitivity in batch}
test2 = analyze_sensitivity(test2,useCurve = 'approach',threads = 1,end = 20)
test2 = analyze_sensitivity(test2,useCurve = 'retract',threads = 1,end = 20)
view_df(test2@metadata,'metadata with sensitivity values added as columns')
```

### Baseline offsets


Similarly, the offset voltage of each curve can be estimated using `find_baseline()`. We assume that the far-right region of every curve corresponds to the probe being fully detached from the surface and therefore free from interaction forces. The mean of the y values of the last `N` points, defined by the `least_length` argument, in this region is calculated as the baseline offset value. This value is then used in later steps to move the curve up or down to align with the zero-force line, that is, the x-axis. Here we set the minimum baseline length to 300, because the heatmap above shows that all curves maintain a stable baseline over the last 300 data points. We set the slope threshold to 0.01 and the standard deviation threshold to 0.01, meaning that a curve will be discarded if the last 300 points have a y-value standard deviation greater than 0.01 or a linear-regression slope greater than 0.01.

Like `calc_sensitivity()`, `find_baseline()` returns two outputs: the offset value and the subset of data points used to calculate it.


```{r single curve baseline calculation}

res.base = find_baseline(x = x, y = y,  sensitivity = res.sens$sensitivity, least_length = 300,slp_threshold = 0.01,std_threshold = 0.01)
example1.raw$`Region used for baseline offset calculation` = case_when(example1.raw$Calc_Ramp_Rt_nm %in% res.base$segment$x ~ "Baseline region", TRUE ~ "Non-baseline region")

ggplot(example1.raw, aes(x = Calc_Ramp_Rt_nm, y = Defl_V_Rt)) +
  geom_point(aes(color = `Region used for baseline offset calculation`), size = 1.5) +
  scale_color_manual(
    values = c(
      "Baseline region" = "orange",
      "Non-baseline region" = "grey70"
    ),
    name = "Region"
  ) +
  labs(
    x = "Calc_Ramp_Ex (nm)",
    y = "Defl_V_Ex (V)",
    title = "AFM Force Curve Regions"
  ) +
  theme_minimal(base_size = 14)
```


To analyze the baseline offset for all curves in an fdObj at once, we use `analyze_baseline()`. This function internally calls `find_baseline()` for each curve and records the computed baseline offset values in the metadata columns `baseline_V_dir`, along with the user-defined length of the baseline segment. The portion of each raw curve used for baseline offset estimation is stored in the object's `baseline_segment` slot.

Curves can exhibit undulating or unstable baselines. Such abnormal curves, identified by a slope greater than `slp_threshold` and a standard deviation greater than `std_threshold`, are considered unreliable and are omitted from subsequent analyses. Curves that fail baseline calculation are assigned `NA` in these metadata columns. `analyze_baseline()` automatically looks up the sensitivity columns generated by `analyze_sensitivity()`. Since baseline estimation requires a valid sensitivity value, any curve that fails sensitivity calculation, that is, does not have a numeric sensitivity value in the corresponding metadata column, will also fail baseline calculation. 

```{r analyze baseline in batch}
test2 = analyze_baseline(test2,useCurve = 'approach',least_length = 300,slp_threshold = 0.01,std_threshold = 0.01)
test2 = analyze_baseline(test2,useCurve = 'retract', least_length = 300,slp_threshold = 0.01,std_threshold = 0.01)
view_df(test2@metadata,'metadata with baseline values added as columns')
```


## Transformation of a raw curve

Now that the sensitivity and baseline offset values have been calculated, we are ready to transform the curve with `transform_a_curve()`. `senscal_seg_x` and `senscal_seg_y` are the x and y values of the segment used for sensitivity calculation, and they are also used for curve transformation. 

```{r single curve transform,fig.height=6,fig.width=16}
example1.transformed = transform_a_curve(x = x, y= y, baseline = res.base$baseline, sensitivity = res.sens$sensitivity,
                             spring_constant = 0.08, senscal_seg_x = res.sens$senscal_segment$x,
                             senscal_seg_y = res.sens$senscal_segment$y)

p1 = ggplot(example1.raw, aes(x = Calc_Ramp_Rt_nm, y = Defl_V_Rt)) +
  geom_point(aes(
    color = case_when(
      `Region used for Sensitivity calculation` == 'Sensitivity region' ~ "Sensitivity region",
      `Region used for baseline offset calculation` == 'Baseline region' ~ "Baseline region",
      TRUE ~ "Other"
    )
  ), size = 1.5) +
  scale_color_manual(
    values = c(
      "Sensitivity region" = "hotpink",
      "Baseline region" = "orange",
      "Other" = "grey70"
    ),
    name = "Region"
  ) +
  labs(
    x = "Calc_Ramp_Ex (nm)",
    y = "Defl_V_Ex (V)",
    title = "Raw curve"
  ) +
  theme_minimal(base_size = 14) + 
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.5)


example1.transformed = cbind(example1.transformed, example1.raw[,c('Region used for Sensitivity calculation','Region used for baseline offset calculation')])
p2 = ggplot(example1.transformed, aes(x = separation_distance_nm, y = force_nN)) +
  geom_point(aes(
    color = case_when(
      `Region used for Sensitivity calculation` == "Sensitivity region" ~ "Sensitivity region",
      `Region used for baseline offset calculation` == "Baseline region" ~ "Baseline region",
      TRUE ~ "Other"
    )
  ), size = 1.5) +
  scale_color_manual(
    values = c(
      "Sensitivity region" = "hotpink",
      "Baseline region" = "orange",
      "Other" = "grey70"
    ),
    name = "Region"
  ) +
  labs(
    x = "Separation distance (nm)",
    y = "force (nN)",
    title = "Transformed curve"
  ) +
  theme_minimal(base_size = 14) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.5)

plot_grid(p1,p2)
```


## Transform curves using the wrapper function `transform_curves()`

At this step, we apply the wrapper function `transform_curves()` to transform all raw curves in the fdObj into calibrated force-distance curves at once. This function streamlines the entire transformation process by internally calling `denoise_curves()`, `analyze_sensitivity()`, and `analyze_baseline()` to denoise raw curves and estimate sensitivity and baseline values. All arguments accepted by these individual functions can be passed directly to `transform_curves()`, allowing full control over the analysis parameters. 

__Note: If the sensitivity and baseline columns already exist in the metadata, the function skips the calculation step and directly uses the existing values for curve transformation. To rerun the calculation, you can remove the existing columns from the metadata.__

To transform a curve, the spring constant of the AFM probe is required to convert deflection (V) to force (nN). `transform_curves()` accepts either a single numeric value, applied to all curves and stored in `metadata$spring_constant`, or a metadata column name containing per-curve spring constants. These values are then used with the computed baseline and sensitivity to generate the final force-distance curves.

```{r transform curves approach,warning=TRUE, message=TRUE,results='hold', echo=TRUE}
t = 1

test = transform_curves(test_beforeTransform,
                        denoise_first = TRUE, # whether to denoise raw curves before transformation. Denoising can improve the reliability of sensitivity and baseline calculations, especially for noisy curves. The smoothed curve will be used for downstream analysis, while the original raw curve will be preserved in the rawCurves slot for reference.
                        p = 1, n = 3, # parameters for the Savitzky-Golay filter used for denoising. p is the polynomial order (default 1, meaning linear), and n is the window size (default 3, meaning each point is smoothed by fitting a polynomial to itself and its 2 neighbors). Users can adjust these parameters based on curve noise and shape.
                        
                        spring_constant = 0.08, # spring constant
                        useCurve = 'approach', # analyze the approach curve

                        # baseline determination and filtering
                        least_length = 300, # use the last 300 points as baseline
                        slp_threshold = 0.01, # any curve with a baseline-region slope > 0.01 will be discarded
                        std_threshold = 0.01, # any curve with a baseline-region standard deviation of y values > 0.01 will be discarded

                        # sensitivity calibration
                        intv = 4, # number of data points added per iteration when building the linear sensitivity segment
                        end = 20, # use at most the first 20 data points on the left side of the curve to calibrate sensitivity
                        R_squared_min = 0.99, # minimum R^2 threshold for accepting a segment as sufficiently linear during sensitivity calculation
                        minimum_length = 4, # minimum number of accumulated points required for a valid sensitivity result; if shorter, sensitivity is reported as NA
                        threads = t)

test = transform_curves(test,
                        denoise_first = TRUE, 
                        p = 1, n = 3, 
                        spring_constant = 0.08, 
                        useCurve = 'retract', # analyze the retract curve

                        # baseline determination and filtering
                        least_length = 300, 
                        slp_threshold = 0.01, 
                        std_threshold = 0.01, 

                        # sensitivity calibration
                        intv = 4, 
                        end = 20, 
                        R_squared_min = 0.99, 
                        minimum_length = 4, 
                        threads = t)
```

## Soft surface correction

__When measuring on a compliant (soft) surface, such as a hydrogel, cell membrane, or polymer brush, the contact region of the curve reflects a combined response from both cantilever deflection and surface indentation. Using this region to estimate probe sensitivity will therefore underestimate the true probe sensitivity. To correct for this, the probe sensitivity should be calibrated separately on a rigid reference surface (for example, bare glass or silicon) and then supplied to `transform_curves()` when processing soft-surface curves.__

__To do this, set `soft = TRUE` and provide the externally calibrated sensitivity via `probe_sensitivity_external`. This can be either a single numeric value, applied to all curves, or the name of a metadata column containing per-curve sensitivity values. When `soft = TRUE`, `transform_curves()` uses this external sensitivity to convert deflection (V) to force (nN). Note that the x-axis transformation, from piezo displacement to separation distance, is unaffected and always uses the sensitivity estimated from each individual curve, preserving the deformation component in the separation distance.__



```{r use external sensitivity value}

test3 = transform_curves(test,
                        useCurve = 'retract', # analyze the retract curve
                        denoise_first = TRUE, 
                        p = 1, n = 3, 
                        spring_constant = 0.08, 
                        # baseline determination and filtering
                        least_length = 300, 
                        std_threshold = 0.01, 
                        slp_threshold = 0.01,
                        # sensitivity calibration
                        intv = 4, 
                        end = 20, 
                        R_squared_min = 0.99, 
                        minimum_length = 4, 
                        threads = 1,

                        # soft surface correction
                        soft = TRUE, # whether to apply soft surface correction
                        probe_sensitivity_external = 0.05 # externally calibrated sensitivity value to use for soft surface correction (in V/nm). This can be a single numeric value applied to all curves, or the name of a metadata column containing per-curve sensitivity values.
                        )
```

## Examine the object after transformation

We can see that the curves have now been transformed.

```{r examine fdobj after transformation}
print(test)
```

The transformed curves are stored in the `retractCurves` and `approachCurves` slots of the fdObj. Note that when sensitivity or baseline cannot be reliably determined, such as in curves with highly fluctuating probe-surface contact regions or irregular baseline segments, the transformation will fail. In these cases, the corresponding entry in the transformed curve slot, `retractCurves` or `approachCurves`, is stored as an empty data frame with zero rows, indicating an unsuccessful transformation.

```{r show retract and approach curve slots}
print(names(test@retractCurves)%>% head()) 
print(names(test@approachCurves)%>% head())
```

The calculated sensitivity, baseline, and input spring constant are now saved as columns in the metadata. Curves with undetermined baseline or sensitivity values will have missing values in the corresponding cells.

```{r show metadata after transformation}
view_df(test@metadata,'metadata after curve transformation')
```

## Plot transformed curves

To visualize the transformed curves, we can use `plot_fd_curves()`. Here, we plot the approach and retract curves in separate panels and color-code them by `region`.

```{r plot transformed curves, fig.height=12,fig.width=10}
plot_fd_curves(test,group_curves_by = 'region', split_curves_by = 'surface',
               point_alpha = 0.5,line_alpha = 0.3, point_size = 0.5,
               color_map = list(`1` = 'darkred',
                                `2` = 'orange',
                                `3` = 'steelblue'))
```

We can see that Group_A has the largest adhesion, which occurs mostly immediately after the probe is pulled off the surface. The other groups also show adhesion, but at lower levels. All curves that passed the transformation step have baselines that are well aligned with the zero-force line and show relatively small fluctuations, which is important for accurate calculation of analytical metrics in later steps.


# Analytical metrics calculation

After curve transformation, we can calculate analytical metrics such as adhesive force, interaction energy, and interaction distance.

## Baseline noise-level determination

### Wrapper function for batch analysis

Before calculating analytical metrics, it is important to determine the baseline noise level so that insignificant fluctuations can be excluded from metric calculations. The function `analyze_a_curve_noise()` estimates lower and upper noise bands from baseline force values. For batch analysis, `analyze_curves_noise()` applies this to all transformed curves in an fdObj and writes the results into metadata columns `noiseBand_low_nN` and `noiseBand_high_nN`, one set for approach curves and one set for retract curves.

Supported noise-band determination methods are standard deviation (`sd`), median absolute deviation (`mad`), quantile (`quantile`), and fixed (`fixed`). The `multiplier` parameter scales the chosen noise metric to define the noise band. The noise bands are then reused in interaction-distance and interaction-energy analyses, so thresholds are determined once and applied consistently across curves.

Here we define the noise band as the range from the minimum to the maximum force in the baseline region.

__Note: Set the noise-band multiplier to 0 to disable noise-band thresholding.__

```{r calculate baseline noise level}
test2 = analyze_curves_noise(test,
                            useCurve = 'approach', # calculate noise band for approach curves,
                            baseline_span = "automatic", # number of data points at the end of the curve to use as baseline region for noise level calculation. Here we set it to "automatic", which means the function will use the baseline_span column in the metadata.
                            threshold_method = 'quantile', # method to determine noise band. Here we use quantile, which means the noise band will be defined by the quantiles of the baseline force values. The specific quantiles are determined by quantile_low and quantile_high parameters.
                            quantile_low = 0, # lower quantile for noise band. Here we set it to 0, meaning the lower bound of the noise band will be the minimum force value in the baseline region.
                            quantile_high = 1, # upper quantile for noise band. Here we set it to 1, meaning the upper bound of the noise band will be the maximum force value in the baseline region.
                            multiplier = 1 # multiplier for the noise metric. We set multiplier to 1 to directly use the min-max range of the force in baseline region without further scaling.
                            ) 
test2 = analyze_curves_noise(test2, useCurve = 'retract',
                            threshold_method = 'quantile',quantile_low = 0,quantile_high = 1,
                            multiplier = 1)
view_df(test2@metadata,'metadata with baseline noise level values added as columns')
```

### Single curve analysis

To demonstrate how the baseline noise level is calculated for a single curve, we use the transformed retract curve above as an example. The function `analyze_a_curve_noise()` estimates the lower and upper noise bands based on the specified method and multiplier, using the baseline region of the curve as discussed above. 

Here we use the same parameters as in the batch analysis, which means the noise band spans the minimum and maximum force values in the baseline region. The function returns a named vector containing the calculated lower and upper noise-band values.

```{r single curve noise level calculation,fig.height=6,fig.width=8}
res.noise = analyze_a_curve_noise(example1.transformed, baseline_span = 300, threshold_method = 'quantile', multiplier = 1,quantile_low = 0,quantile_high = 1)
print(res.noise)

```


## Adhesive force

### Wrapper function for batch analysis

The adhesive force is calculated as the most negative force value in a curve, which represents the maximum adhesive force between the AFM probe and the sample surface. `analyze_curves_adhesive_force()` identifies this value and the separation distance at which it occurs, and records the force in the metadata column `adhesive_force_nN` and the distance in the column `adhesive_sep_nm`. Curves that fail transformation will have `NA` in these columns.

__Note: If the minimum force of the curve is positive, the adhesive force is reported as 0 and the associated separation distance is reported as `NA`. If the minimum force is negative, the adhesive force is reported as the absolute value of that force.__

```{r calculate adhesive force}
test2 = analyze_curves_adhesive_force(test2,useCurve = 'approach',threads = 1)
test2 = analyze_curves_adhesive_force(test2,useCurve = 'retract',threads = 1)
view_df(test2@metadata,'metadata with adhesive force values added as columns')   
```

### Single curve analysis

The function `analyze_a_curve_adhesive_force()` identifies the lowest negative force value in the curve and the corresponding separation distance, and returns the absolute value of the force and the separation distance in a named vector.

```{r single curve adhesive force calculation}
res.af = analyze_a_curve_adhesive_force(example1.transformed)
print(res.af)
```

## Interaction energy

In AFM force spectroscopy, interaction energy is quantified by calculating the area under the curve. `analyze_a_curve_area()` computes adhesive energy, the area below the negative noise threshold and above the curve in Quadrant IV, representing bond-breaking energy, and repulsive energy, the area above the positive noise threshold and under the curve in Quadrant I. A noise band is applied to filter out thermal fluctuations and instrumental noise, which would otherwise artificially inflate energy measurements. By excluding force values within the noise band, the function ensures that only biologically or physically meaningful interactions are counted, yielding more accurate energy estimates.

### Wrapper function for batch analysis

The wrapper function `analyze_curves_energy()` applies this analysis to all curves in an fdObj and records the calculated adhesive and repulsive energy in the metadata, in the columns `adhesive_energy_fJ` and `repulsive_energy_fJ`, with one set for retract curves and one set for approach curves. Curves that fail transformation will have `NA` in these columns. The function automatically looks up the pre-determined noise band in the object's metadata.

Note that for over-corrected curves, in which part of the curve near the contact region extends to the negative side of the x-axis, that is, has negative separation distance, the function automatically sets all negative x values to zero before energy calculation.

```{r calculate interaction energy}
test2 = analyze_curves_energy(test2,useCurve = 'approach',threads = 1)
test2 = analyze_curves_energy(test2,useCurve = 'retract',threads = 1)

view_df(test@metadata,'metadata with energy values added as columns') 
```

### Single curve analysis

`analyze_a_curve_area()` computes adhesive and repulsive energy from the area under the curve while applying a noise-band to exclude insignificant fluctuations. To avoid counting noise as part of the interaction energy, we set the noise cutoff using the computed noise band. 

```{r single curve interaction energy calculation, fig.height=6, fig.width=8}
analyze_a_curve_area(example1.transformed, noiseBand_low = res.noise[['noiseBand_low']], noiseBand_high = res.noise[['noiseBand_high']])
```

## Interaction distance

Interaction distance is calculated as the separation distance at which the force curve first crosses the noise threshold, indicating the start or end of significant interaction between the AFM probe and the sample surface. By configuring the function to scan in different directions, left-to-right or right-to-left, and to look for either positive or negative excursions, users can identify different types of interaction distances. 


### Rupture/adhesion distance

For rupture events, which are in negative-force excursions, the function can be set to scan from right to left and identify the first point where the curve dips below the negative noise threshold, which in turn indicates the end of significant adhesion.

Specifically, the function scans the curve from right to left, `x_direction = "left"`, while excluding the baseline region defined by `baseline_span`, to look for the first negative-force excursion, `y_direction = "negative"`, with at least 3 consecutive points below the threshold, `min_consecutive = 3`.

```{r single curve rupture distance}
res.rupturedis = analyze_a_curve_interaction_distance(example1.transformed,
                                                      y_direction = "negative", 
                                                      x_direction = "left",
                                                      baseline_span = 200,
                                                      noiseBand_low = res.noise[['noiseBand_low']],
                                                      noiseBand_high = res.noise[['noiseBand_high']],
                                                      min_consecutive = 3
                                                      )

print(res.rupturedis)

```

To analyze rupture distance for all curves in an fdObj, we can use the wrapper function `analyze_curves_interaction_distance()` with the same parameters. The calculated rupture distance is recorded in a metadata column named `rupture_distance_nm`. Curves that fail transformation will have `NA` in this column.

```{r rupture distance for fdobj}
test2 = analyze_curves_interaction_distance(test2,
                                           useCurve = 'retract', # analyze retract curve
                                           y_direction = "negative", x_direction = "left",
                                           baseline_span = 'automatic', # use the baseline span values in metadata column "baseline_span_retract" for each curve
                                            min_consecutive = 3
                                            )

view_df(test2@metadata,'metadata with rupture length')
```

### Repulsive distance

For repulsive interactions, which are positive-force excursions, the function can be set to scan from left to right and identify the last point before the curve crosses the positive noise threshold and enters the noise band from the repulsive region, which marks the end of significant repulsion.

Specifically, the function scans the curve from left to right, `x_direction = "right"`, up to the beginning of the baseline region, as defined by `baseline_span`. It identifies the last data point where the curve first crosses the positive noise threshold, `y_direction = "positive"`, entering the noise band from the repulsive-force region. Because the curve is expected to begin in the repulsive region, `min_consecutive` is set to 1, meaning that a single point above the positive noise threshold is sufficient to mark the start of the repulsive region.


```{r single curve repulsive distance}
res.repuldis = analyze_a_curve_interaction_distance(example1.transformed,
                                                      y_direction = "positive", 
                                                      x_direction = "right",
                                                      baseline_span = 200,
                                                      noiseBand_low = res.noise[['noiseBand_low']],
                                                      noiseBand_high = res.noise[['noiseBand_high']],
                                                      min_consecutive = 1
                                                      )

print(res.repuldis)

```

To analyze repulsive distance for all curves in an fdObj, we can use the wrapper function `analyze_curves_interaction_distance()` with the same parameters. The calculated repulsive distance is recorded in a metadata column named `repulsive_distance_nm`. Curves that fail transformation will have `NA` in this column.

```{r repulsive distance for fdobj}
test2 = analyze_curves_interaction_distance(test2,
                                           useCurve = 'approach', # analyze approach curve
                                           y_direction = "positive", x_direction = "right",
                                           baseline_span = 'automatic', # use the baseline span values in metadata column "baseline_span_approach" for each curve
                                           min_consecutive = 1
                                          )

view_df(test2@metadata,'metadata with repulsive distance')
```

## All-in-one analytical wrapper

You can run all analytical-metric steps described above in one call using `analyze_curves_all_analytical_metrics()`. This function combines the analyses above into a single streamlined workflow, allowing you to compute noise levels, adhesive forces, interaction energies, and interaction distances for all curves in an fdObj with one command. By setting the appropriate toggles, you can customize which metrics to calculate and how to apply noise thresholds, providing a comprehensive analysis of your AFM force spectroscopy data in a single step.

```{r run all analytical metrics with toggles}
test = analyze_curves_all_analytical_metrics(
  test, # the object
  useCurve = "both", # run analysis for both approach and retract 
  threads = 1, 

  # noise band parameters
  noise_baseline_span = "automatic", # use the baseline section previously defined in curve transformation for noise band calculation
  noise_threshold_method = "quantile", # use quantile method to define the noise band
  noise_quantile_low = 0.00, # define the lower end of noise band as minimum y value of the baseline section
  noise_quantile_high = 1, # define the upper end of noise band as maximum y value of the baseline section
  noise_multiplier = 1, # multiply the noise band by 1 to directly use the min-max range of the force in baseline region without further scaling.

  # adhesive force
  analyze_adhesive_force = TRUE, # whether to analyze adhesive force

  # adhesive energy and repulsive energy
  analyze_energy = TRUE, # whether to analyze interaction energy (repulsive and adhesive)

  # rupture (adhesive) distance
  analyze_rupture_distance = TRUE, # whether to analyze adhesive/rupture distance
  analyze_rupture_distance_baseline_span = "automatic", # use the baseline section previously defined in curve transformation for rupture distance analysis
  analyze_rupture_distance_x_direction = "left", # scan from right to left to find the first data point at which a curve enters the adhesive region from the noise band region
  analyze_rupture_distance_min_consecutive = 3, # to reduce false positives from random fluctuations, require at least 3 consecutive points below the lower noise-band threshold for classifying the curve as entering the adhesive region.

  # repulsive distance
  analyze_repulsive_distance = TRUE, # whether to analyze repulsive distance
  analyze_repulsive_distance_baseline_span = "automatic", # use the baseline section previously defined in curve transformation for repulsive distance analysis
  analyze_repulsive_distance_x_direction = "right", # scan from left to right to find the last data point before a curve first enters the noise band region from the repulsive region
  analyze_repulsive_distance_min_consecutive = 1 # because the curve is expected to start from the repulsive region, we set min_consecutive to 1, meaning that a single point above the positive noise threshold is sufficient to mark the start of the repulsive region.
)
```


# Visualization of analytical metrics

`plot_a_curve_metrics()` can be used to visualize the calculated analytical metrics for a single curve.

```{r single curve plot metrics,fig.height=12, fig.width=22}
plot_a_curve_metrics(
  fdobj = test,
  curve_name = example1.name,
  useCurve = "retract", # which curve to plot, can be "approach" or "retract".
  plot_raw = T, # whether to plot the raw curve together with the transformed curve. 
  base_size = 15,
  annotation_text_size = 5
)

```

# Save all raw and transformed curves with analytical metrics annotated (optional)

```{r plot all metrics for a curve,fig.height=8, fig.width=16,eval=F}
plot_all_curve_metrics(
  fdobj = test, # the fdObj to plot
  useCurve = "retract", # plot the retract curve
  plot_raw = T, # whether to plot the raw curve
  destination_folder = '../test_plot_metrics_denoised', # the path to save the plots
  format = 'png', width = 18, height = 9, # plot format and dimensions,
  threads = 8 # number of threads to use for parallel processing (set to 8 to use 8 cores, which will significantly speed up the plotting process for a large dataset
)
plot_all_curve_metrics(
  fdobj = test, # the fdObj to plot
  useCurve = "approach", # plot the approach curve
  plot_raw = T, # whether to plot the raw curve
  destination_folder = '../test_plot_metrics_denoised', # the path to save the plots
  format = 'png', width = 18, height = 9, # plot format and dimensions
  threads = 8 # number of threads to use for parallel processing (set to 8 to use 8 cores, which will significantly speed up the plotting process for a large dataset
)
```

# Summary statistics of analytical metrics

## PCA biplot of selected metrics

Dimensionality reduction is first used to visualize the overall pattern of how curves cluster based on their analytical metrics. The function `plot_pca_biplot()` performs principal component analysis (PCA) on the selected metrics and plots sample scores in PCA space, colored by a specified metadata variable such as surface type. The function also plots feature loadings as arrows, indicating how each metric contributes to the principal components and the clustering pattern of samples.

```{r pca biplot of metrics, fig.height=6, fig.width=7}
pca_features <- c(
  "adhesive_force_nN_retract",
  "adhesive_energy_aJ_retract",
  "rupture_distance_nm_retract",
  "repulsive_energy_aJ_approach",
  "repulsive_distance_nm_approach"
)

plot_pca_biplot(
  df = test@metadata,
  include_columns = pca_features,
  color_by = "surface",
  color_map = c(
    Group_B = "#1b9e77",
    Group_C = "#d95f02",
    Group_A = "#7570b3",
    Group_D = "#e7298a"
  ),
  arrow_scale = 1.1,
  show_feature_labels = TRUE
)
```

From the PCA biplot, we can see that Group_A is well separated from the other three groups, which is consistent with our observations of the curves and with the fact that Group_A has the largest adhesion. The biplot also shows that adhesive force, adhesive energy, and rupture distance are the main contributors to PC1, which separates Group_A from the others, while repulsive energy and repulsive distance contribute more to PC2, which largely separates Group_D from Group_B and Group_C. Adhesive force, adhesive energy, and rupture distance are positively correlated with one another, which is consistent with the fact that they all reflect adhesion strength. Repulsive energy and repulsive distance are also positively correlated, which makes sense because a stronger repulsion would typically extend over a longer distance.


## Scaled feature heatmap with sample annotations

The function `plot_complex_heatmap()` creates a heatmap of selected analytical metrics across samples, with options for clustering and annotation. In this example, we select the same metrics used in the PCA plot to visualize their relative values across samples. The heatmap is scaled by z-score to show how each sample deviates from the mean for each metric, allowing easy comparison across metrics with different units and ranges. Samples are annotated by surface type and region, with a specified color map for the surface types. Clustering is applied to rows (samples) to reveal patterns of similarity based on the selected metrics.

```{r complex heatmap of metrics, fig.height=5, fig.width=8}
  plot_complex_heatmap(
    df = test@metadata,
    include_columns = pca_features,
    annotate_columns = c("surface", "region"),
    annotation_colors = list(
      surface = c(
        Group_B = "#1b9e77",
        Group_C = "#d95f02",
        Group_A = "#7570b3",
        Group_D = "#e7298a"
      )
    ),
    cluster_rows = TRUE,
    cluster_columns = FALSE,
    show_row_dend = TRUE,
    show_column_dend = FALSE,
    heatmap_name = "z-score",
    show_column_names = FALSE
  )
```

## Violin plot of selected metrics by surface

`plot_metric_violin()` can be used to visualize the distribution of selected metrics across different groups. Here, we plot violin plots for adhesive force, adhesive energy, and rupture distance from the retract curves, as well as repulsive energy and repulsive distance from the approach curves. The option `show_whisker_box = TRUE` displays the median and interquartile range of each distribution.

```{r summary statistics of metrics, fig.height=6, fig.width=12}
f1 = plot_metric_violin(test@metadata, metric_name = "adhesive_force_nN_retract", group_by = "surface", color_map = c(Group_B = "#1b9e77", Group_C = "#d95f02", Group_A = "#7570b3", Group_D = "#e7298a"),
show_whisker_box = T)
f2 = plot_metric_violin(test@metadata, metric_name = "adhesive_energy_aJ_retract",log10=T,  group_by = "surface", color_map = c(Group_B = "#1b9e77", Group_C = "#d95f02", Group_A = "#7570b3", Group_D = "#e7298a"),
show_whisker_box = T)
f3 = plot_metric_violin(test@metadata, metric_name = "repulsive_energy_aJ_approach", group_by = "surface", color_map = c(Group_B = "#1b9e77", Group_C = "#d95f02", Group_A = "#7570b3", Group_D = "#e7298a"),
show_whisker_box = T)
f4 = plot_metric_violin(test@metadata, metric_name = "rupture_distance_nm_retract", log10=T, group_by = "surface", color_map = c(Group_B = "#1b9e77", Group_C = "#d95f02", Group_A = "#7570b3", Group_D = "#e7298a"),
show_whisker_box = T)
f5 = plot_metric_violin(test@metadata, metric_name = "repulsive_distance_nm_approach", log10=T, group_by = "surface", color_map = c(Group_B = "#1b9e77", Group_C = "#d95f02", Group_A = "#7570b3", Group_D = "#e7298a"),
show_whisker_box = T)

plot_grid(f1,f2,f3,f4,f5, nrow = 2)


```

## Violin plot with statistical annotations

`plot_metric_violin()` also has options to perform statistical tests and annotate significance on the plot. Here, we perform a Kruskal-Wallis test to compare adhesive force across different groups, followed by pairwise Wilcoxon tests for post hoc analysis. The p-values are adjusted using the Benjamini-Hochberg method to control the false discovery rate.

```{r summary statistics with tests, eval=requireNamespace("ggpubr", quietly = TRUE)}
plot_metric_violin(
  df = test@metadata,
  metric_name = "adhesive_force_nN_retract",
  group_by = "surface",
  color_by = "surface",
  color_map = c(
    Group_B = "#1b9e77",
    Group_C = "#d95f02",
    Group_A = "#7570b3",
    Group_D = "#e7298a"
  ),
  global_test = "kruskal",
  pairwise_test = "wilcox",
  p_adjust_method = "BH"

)
```

The function also allows users to specify which pairwise comparisons to perform so that only relevant comparisons are made and the plot is not overcrowded with significance annotations. Here, we compare Group_A with each of the other three groups.

```{r summary statistics with tests2, eval=requireNamespace("ggpubr", quietly = TRUE)}
plot_metric_violin(
  df = test@metadata,
  metric_name = "adhesive_force_nN_retract",
  group_by = "surface",
  color_by = "surface",
  color_map = c(
    Group_B = "#1b9e77",
    Group_C = "#d95f02",
    Group_A = "#7570b3",
    Group_D = "#e7298a"
  ),
  global_test = "none", # do not perform global test, only do pairwise comparisons
  pairwise_test = "wilcox",
  p_adjust_method = "BH", pairwise_comparisons = list(
    c("Group_A", "Group_D"),
    c("Group_A", "Group_C"),
    c("Group_A", "Group_B")
  )
)
```
